---
title: "From a RWE-BRIDGE codelist to a Synthea module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a RWE-BRIDGE codelist to a Synthea module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")` and `vignette("cohort-building")`. Those
cover authoring a module's *content* directly in R. This one covers a different starting point:
you already have a **RWE-BRIDGE-style reference codelist** -- a CSV mapping concepts to real
codes, of the `event_definition / coding_system / code / concept_name / tags / type /
event_abbreviation` shape used across several UMC-Utrecht-RWE study repositories -- and want a
Synthea module that emits raw exposure/outcome records using those real codes.

## The four stages

```{r setup}
library(syntheaModuleR)
```

1. **Specify where the necessary information is** -- `read_bridge_codelist()` reads a codelist
   file (or an already-loaded `data.frame`) and returns a clean, canonically-named code table.
2. **Add inclusion criteria / exposure / outcome parameters** -- `build_module_spec()` combines
   those resolved codes with the parameters a codelist alone can never supply (arm shares, onset
   probability/timing, optional demographic eligibility).
3. **Produce a YAML that is editable** -- `write_module_spec()` persists the result; every code,
   share, and probability is right there in plain text, ready to hand-edit.
4. **Go from YAML to build a module** -- `build_module_from_spec()` reads it back and mechanically
   assembles the Synthea module. This step never touches a codelist again.

## Stage 1: read the codelist

A toy codelist, standing in for a real RWE-BRIDGE file (the concept `GRIPPE` deliberately has one
`exclude`-tagged and one `narrow`-tagged row, to show the tag-preference filtering):

```{r toy-codelist}
codelist <- data.frame(
  event_abbreviation = c("GRIPPE", "GRIPPE", "COUGH"),
  coding_system = "SNOMED-CT",
  code = c("1", "2", "3"),
  code_name = c("An unrelated excluded concept", "Influenza (disorder)", "Cough (finding)"),
  tags = c("exclude", "narrow", "narrow"),
  type = "AESI",
  stringsAsFactors = FALSE
)

outcome_codes <- read_bridge_codelist(
  codelist,
  type = "AESI", coding_system = "SNOMED-CT",
  group_by = "event_abbreviation"
)
outcome_codes
```

The `exclude`-tagged row for `GRIPPE` was never a candidate -- `read_bridge_codelist()` picked the
`narrow` one. A real RWE-BRIDGE codelist often has dozens of candidate rows per concept; this is
what makes picking the right one by hand tedious enough to be worth automating (and, done
carelessly, easy to get wrong -- an earlier, unfiltered extraction genuinely picked an
`exclude`-tagged row for a real AESI).

An exposure-style extraction looks different: a handful of already-distinct rows (e.g. specific
vaccine products), filtered by `event_abbreviation` alone, with no tag ranking needed
(`group_by = NULL`, the default):

```{r toy-exposure}
drug_codelist <- data.frame(
  event_abbreviation = c("FLU_VACC", "FLU_VACC"),
  product_identifier = "PRODCODEID",
  code = c("101", "102"),
  product_name = c("Brand A flu vaccine", "Brand B flu vaccine"),
  stringsAsFactors = FALSE
)

flu_rows <- read_bridge_codelist(
  drug_codelist,
  event_abbreviation = "FLU_VACC", coding_system = "PRODCODEID",
  columns = bridge_columns(coding_system = "product_identifier", display = "product_name")
)
flu_rows
```

```{r toy-exposure-codes}
exposure_codes <- data.frame(
  option = c("brand_a", "brand_b"),
  system = "PRODCODEID",
  code = flu_rows$code,
  display = flu_rows$display,
  stringsAsFactors = FALSE
)
```

## Stage 2: add the parameters a codelist can't supply

```{r spec}
spec <- build_module_spec(
  "flu_vaccine_example",
  exposure_codes = exposure_codes,
  exposure_shares = c(brand_a = 0.3, brand_b = 0.2, comparator = 0.5),
  outcome_codes = outcome_codes,
  outcome_probability = 0.05,
  outcome_delay = list(low = 0, high = 365, unit = "days"),
  inclusion_criteria = list(age = list(operator = ">=", quantity = 18, unit = "years"))
)
```

Note what's *not* here: no eligibility window, no composite subgroup logic, no incidence rate
sourced from any pipeline-computed file. `inclusion_criteria` is the one optional exception --
plain demographic skew, off by default, and never derived automatically.

## Stage 3: write (and read back) the YAML

```{r yaml, comment = ""}
path <- tempfile(fileext = ".yaml")
write_module_spec(spec, path)
cat(readLines(path), sep = "\n")
```

This file is the thing you'd commit and hand-edit -- fix a code, add or remove an outcome item,
adjust a share, all without touching R again.

## Stage 4: build the module

```{r build}
module_json <- build_module_from_spec(path)
parsed <- jsonlite::fromJSON(module_json, simplifyVector = FALSE)
length(parsed$states)
parsed$states$Initial$direct_transition
```

`inclusion_criteria` was given, so `Initial` routes into a `create_population()` Guard first; omit
it entirely (`inclusion_criteria = NULL`, the default) and `Initial` would route straight into the
exposure block instead -- `chain()` drops a `NULL` fragment on its own, no special-casing needed.

