---
title: "From a RWE-BRIDGE codelist to Synthea modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a RWE-BRIDGE codelist to Synthea modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")` and `vignette("cohort-building")`. Those
cover authoring a module's *content* directly in R. This one starts somewhere else. You already
have one or more **RWE-BRIDGE-style reference codelists**: CSVs mapping study concepts (outcomes,
covariates, drug classes, ...) to real codes, in the `event_definition / coding_system / code /
concept_name / tags / type / event_abbreviation` shape used across several UMC-Utrecht-RWE study
repositories. You want synthetic patients whose records use those real codes: an exposure, prior
history before it, and outcomes after it, so a study pipeline that runs on the codelists has
something to find for every concept.

```{r setup}
library(syntheaModuleR)
```

## The four stages

1. **Read every concept and code**: `read_bridge_concepts()` returns one row per (concept, code)
   for every eligible row of a codelist.
2. **Add the study parameters**: `build_event_spec()` assigns each concept to a *block* and a
   *pattern* and adds the parameters a codelist can't supply: exposure arms, the timeline, and
   how many events a patient gets.
3. **Write an editable spec**: `write_event_spec()` writes a YAML (structure and parameters) plus
   two CSVs (concepts, codes). These are the files you review, commit and hand-edit.
   `read_event_spec()` reads them back and checks them.
4. **Build the modules**: `build_modules_from_events()` mechanically turns the spec into a few
   Synthea modules, and `write_modules()` writes them where Synthea can run them.

Nothing is read from a study's pipeline-computed logic (e.g. `study_variables.csv`); every
parameter is supplied by the caller.

## Stage 1: read every concept and code

Two toy codelists stand in for the real files. The first has the full-codelist shape (concept id
in `variable_name`, category in `type`); the second has the drug-proxy shape (concept id in
`drug_abbreviation`, category in `system`, product columns instead of `coding_system`/`code_name`):

```{r toy-codelists}
full <- data.frame(
  variable_name = c("N_GBS_AESI", "N_GBS_AESI", "N_GBS_AESI", "C_HF_COV", "C_HF_COV"),
  event_abbreviation = c("GBS", "GBS", "GBS", "HF", "HF"),
  type = c("AESI", "AESI", "AESI", "COV", "COV"),
  coding_system = "SNOMEDCT_US",
  code = c("40956001", "1", "2", "84114007", "42343007"),
  code_name = c(
    "Guillain-Barré syndrome", "Unrelated", "Miller Fisher syndrome",
    "Heart failure", "Congestive heart failure"
  ),
  tags = c("narrow", "exclude", "possible", "narrow", ""),
  stringsAsFactors = FALSE
)

drugs <- data.frame(
  drug_abbreviation = c("DP_STAT", "DP_STAT", "DP_ANTIBIO", "VP_FLU", "VP_FLU"),
  system = c("DP", "DP", "DP", "VP", "VP"),
  product_identifier = "RxNorm",
  code = c("617312", "617311", "308182", "brand_a", "brand_b"),
  product_name = c(
    "Atorvastatin 40 MG", "Atorvastatin 20 MG", "Amoxicillin 250 MG",
    "Brand A flu vaccine", "Brand B flu vaccine"
  ),
  tags = "narrow",
  stringsAsFactors = FALSE
)
```

```{r read}
full_codes <- read_bridge_concepts(full, coding_system = "SNOMEDCT_US")
full_codes

drug_codes <- read_bridge_concepts(
  drugs,
  coding_system = "RxNorm",
  concept_col = "drug_abbreviation", source_type_col = "system",
  fallback_concept = NULL, exclude_concepts = "VP_FLU",
  columns = bridge_columns(coding_system = "product_identifier", display = "product_name")
)
drug_codes
```

A few things happened along the way:

- The `exclude`-tagged GBS row was dropped. `possible`-tagged and blank-tagged rows were kept,
  because by default a row is eligible unless it's tagged `exclude`/`ignore`.
- Every eligible code was kept, not just one per concept. A patient who gets a concept gets one
  of its codes at random, so across a population every code can show up.
- `VP_FLU` was excluded because it's going to be the exposure (below). A concept with no code in
  the requested coding system would be listed in `attr(x, "skipped")` instead of causing an
  error.

The exposure products are picked out individually, as before:

```{r exposure-codes}
flu <- read_bridge_codelist(
  drugs,
  event_abbreviation = "VP_FLU", coding_system = "RxNorm",
  columns = bridge_columns(
    event_abbreviation = "drug_abbreviation",
    coding_system = "product_identifier", display = "product_name"
  )
)
exposure_codes <- data.frame(
  option = c("brand_a", "brand_b"),
  system = "RxNorm",
  code = flu$code,
  display = flu$display,
  stringsAsFactors = FALSE
)
```

## Stage 2: add the study parameters

```{r spec}
x <- build_event_spec(
  "flu_vaccine_example",
  codes = rbind(full_codes, drug_codes),
  exposure_codes = exposure_codes,
  exposure_shares = c(brand_a = 0.3, brand_b = 0.2, comparator = 0.5),
  exposure_delay = list(low = 365, high = 545, unit = "days"),
  cohort_criteria = list(age = list(operator = ">=", quantity = 60, unit = "years")),
  recurring_drugs = "DP_STAT"
)
x$concepts
```

### The timeline

```
cohort entry ──[ baseline window: history ]──────┬── exposure (365–545 days later)
       │                                         │
       └──[ chronic: repeat prescriptions ......................... ]
                                                 └──[ follow-up window: outcomes ]
```

- **Cohort entry.** The exposure module waits for `cohort_criteria` (age >= 60 here), sets the
  `cohort_entry` attribute, waits `exposure_delay`, then assigns the arm. The random delay
  spreads exposure dates instead of everyone being exposed the moment they qualify.
- **Before exposure.** The `baseline` and `chronic` blocks start at cohort entry. They are
  checked so that baseline records always land before the earliest possible exposure, and
  recurring therapy always runs past the latest possible exposure (so both short and long
  look-backs from exposure find prescriptions, in every arm).
- **After exposure.** The `follow_up` block starts when the exposure attribute is set.

### Blocks: how many events a patient gets

Each block becomes **one** module that draws `k` events per patient. Each draw picks a concept
(by `weight`), then one of that concept's codes (by `weight`):

```{r blocks}
str(x$spec$blocks$baseline)
```

With these defaults, 40 % of patients have no baseline history, and the rest have 1 to 5 events.
There are a few modules per study, not one per concept. Synthea re-checks every module for every
person at every timestep, so hundreds of per-concept modules would mostly sit idle in a Guard
for decades of simulated time.

### Patterns: what one event looks like

```{r patterns}
str(x$spec$patterns)
```

- `chronic` conditions are never ended; `acute` conditions are ended after `resolves_after`.
- `course` drugs are a single prescription ended after `course`.
- `recurring` drugs repeat `course` + `gap` for `cycles` prescriptions.
- `once` (vaccines) has no parameters.

Ending a record matters: Synthea doesn't create a new record for a code that's still active.

`build_event_spec()` assigns the defaults per codelist category: AESIs both as history
(`baseline`) and as outcomes (`follow_up`), COV concepts as chronic baseline conditions, drugs as
courses (or `recurring`, in the `chronic` block, if listed in `recurring_drugs`), and vaccines as
`once`. See `default_event_roles()`.

### What a patient will get

`summarise_events()` gives the review summary:

```{r summary}
s <- summarise_events(x)
s$blocks
s$concepts
```

It also reports `s$overlap`: concepts that share codes. A shared code is recorded once but counts
toward every concept that lists it, so overlapping concepts end up more frequent than their
weight alone suggests.

## Stage 3: write (and read back) the spec

```{r write-spec}
dir <- tempfile("spec")
path <- write_event_spec(x, dir)
list.files(dir)
```

```{r yaml, comment = ""}
cat(readLines(path, encoding = "UTF-8"), sep = "\n")
```

The YAML holds the structure. The two CSVs hold everything per concept and per code. In
`concepts.csv` you can:

- change a concept's `weight` (blank means 1);
- set `enabled` to `FALSE` to leave a concept out;
- move a concept to another `block` or `pattern`;
- write the reasoning in `note`.

`read_event_spec()` rejects unknown keys or columns (a typo can't be silently ignored), patterns a
state type can't have, `k` shares that don't sum to 1, and timelines that break the
before/after-exposure guarantees.

## Stage 4: build and write the modules

```{r build}
modules <- build_modules_from_events(path)
names(modules)
```

The `chronic` block has `parallel = 2`. Draws inside one module happen one after another, so a
second long-term therapy in the same module would only start after the first one ended. Two
independent copies let a patient be on up to two therapies at once.

The block modules are built directly on the Engine layer: a K-draw loop whose leaves are one state
per code, all converging on a few shared tail states.

```{r loop}
baseline <- jsonlite::fromJSON(modules[["flu_vaccine_example - baseline"]], simplifyVector = FALSE)
names(baseline$states)
```

Conditions are recorded with no target encounter, so Synthea records them at the exact simulated
time instead of waiting for the next wellness visit.

`write_modules()` writes the files into one flat folder and returns the matching Synthea
arguments. Synthea treats JSON files in a *subfolder* of a module directory as submodules, which
never run on their own.

```{r write, eval = FALSE}
out <- write_modules(modules, "modules/flu_vaccine_example")
out$synthea_args
#> [1] "-d" "modules/flu_vaccine_example" "-m" "flu_vaccine_example*"
```

For quick runs, `build_modules_from_events(path, max_codes_per_concept = 5)` keeps a reproducible
sample of at most 5 codes per concept.

## The original exposure + outcomes spec

`build_module_spec()` / `build_module_from_spec()` (one best code per outcome, one module per
outcome, no records before exposure) are kept unchanged for existing specs, but the events
pipeline above supersedes them.
