---
title: "Building Cohorts Incrementally: Diagnosis, Treatment Pathways, and Switching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Cohorts Incrementally: Diagnosis, Treatment Pathways, and Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")` and `vignette("logic-and-guards")`. Every
other vignette builds one concept in isolation. This one instead walks through a realistic
*workflow*: growing a single module, one `states[["Name"]] <- ...` assignment at a time, until it
produces a population with the kind of heterogeneity a real cohort study needs -- some patients on
one first-line drug, some on another, some on none at all, and a subset who switch therapy or stop
it altogether partway through, without losing track of the patients who didn't.

Concretely, we'll build a module where simulated patients:

1. get diagnosed with type 2 diabetes,
2. split across two first-line drugs -- or get no medication at all,
3. some fraction later switch therapy -- to the other first-line drug, to a third, or discontinue
   treatment entirely,
4. and are finally classified as "treated" or "untreated" by reading their actual medication
   record, not by which pathway through the graph they took to get there.

The example medical codes below (RxNorm/SNOMED-CT) are plausible placeholders chosen for realism
-- verify against current code sets before using this for anything beyond synthetic-data
generation.

```{r setup}
library(syntheaModuleR)

direct <- function(to) create_transition_settings("direct", to = to)

diabetes_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "44054006", display = "Diabetes mellitus type 2 (disorder)"
)
metformin_code <- create_component_settings("code",
  system = "RxNorm", code = "860975", display = "Metformin hydrochloride 500 MG Oral Tablet"
)
sulfonylurea_code <- create_component_settings("code",
  system = "RxNorm", code = "310488", display = "Glipizide 5 MG Oral Tablet"
)
third_agent_code <- create_component_settings("code",
  system = "RxNorm", code = "1545653", display = "Empagliflozin 25 MG Oral Tablet"
)
```

`direct()` is a one-line helper to cut down on repeating
`create_transition_settings("direct", to = ...)` -- nothing more than that.

## Step 1: the base cohort -- diagnosis only

Start as small as possible: everyone in the cohort gets diagnosed with diabetes at a wellness
visit and nothing else happens. This is already a complete, valid module on its own -- the point
of building incrementally is that every step along the way is something you could hand to
`build_module()` and get back a module that loads.

```{r step1}
states_v1 <- list(
  Initial = create_state_settings("Initial", transition = direct("Diabetes Onset")),
  "Diabetes Onset" = create_state_settings("ConditionOnset",
    codes = list(diabetes_code), target_encounter = "Diagnosis Encounter",
    transition = direct("Diagnosis Encounter")
  ),
  "Diagnosis Encounter" = create_state_settings("Encounter",
    wellness = TRUE, transition = direct("End Diagnosis Encounter")
  ),
  "End Diagnosis Encounter" = create_state_settings("EncounterEnd",
    transition = direct("Terminal")
  ),
  Terminal = create_state_settings("Terminal")
)

cat(build_module(name = "Cohort Building Vignette", states = states_v1))
```

Every patient who reaches `Terminal` here has exactly one thing in common: an active diabetes
diagnosis. No medication, no subgroups yet -- that's what the next steps add, without touching
anything already written above.

## Step 2: two first-line drugs, and a group that gets neither

Redirect `End Diagnosis Encounter` away from `Terminal` and into a `distributed_transition` that
splits the cohort three ways: metformin, a sulfonylurea, or no medication at all. Each pathway
tags a `diabetes_regimen` attribute so later states (and, eventually, whoever analyzes the
generated data) can tell which path a given patient took.

```{r step2}
states_v2 <- states_v1

states_v2[["End Diagnosis Encounter"]] <- create_state_settings("EncounterEnd",
  transition = direct("First-Line Treatment Choice")
)

states_v2[["First-Line Treatment Choice"]] <- create_state_settings("Simple",
  transition = create_transition_settings("distributed", options = list(
    list(transition = "Start Metformin", distribution = 0.5),
    list(transition = "Start Sulfonylurea", distribution = 0.3),
    list(transition = "No Medication", distribution = 0.2)
  ))
)

states_v2[["Start Metformin"]] <- create_state_settings("MedicationOrder",
  codes = list(metformin_code), reason = "Diabetes Onset", chronic = TRUE,
  transition = direct("Tag Metformin Regimen")
)
states_v2[["Tag Metformin Regimen"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "metformin", transition = direct("Terminal")
)

states_v2[["Start Sulfonylurea"]] <- create_state_settings("MedicationOrder",
  codes = list(sulfonylurea_code), reason = "Diabetes Onset", chronic = TRUE,
  transition = direct("Tag Sulfonylurea Regimen")
)
states_v2[["Tag Sulfonylurea Regimen"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "sulfonylurea", transition = direct("Terminal")
)

states_v2[["No Medication"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "none", transition = direct("Terminal")
)

cat(build_module(name = "Cohort Building Vignette", states = states_v2))
```

Note what `"No Medication"` *doesn't* do: no `MedicationOrder` state at all. "Some patients take
no medication" isn't a flag you set afterwards -- it's a pathway that never issues an order in the
first place. `reason = "Diabetes Onset"` on the two `MedicationOrder` states resolves against the
`ConditionOnset` state of that name from step 1, per the field's resolution rule
(`vignette("state-reference")`).

## Step 3: some patients switch, and some discontinue entirely

Now make a slice of the metformin pathway revisit their treatment after a delay: most stay on
metformin, some switch to the sulfonylurea, some switch to a third agent entirely -- "another
medication" that was never a first-line option at all -- and some discontinue diabetes medication
altogether, with nothing started in its place. The sulfonylurea and no-medication pathways from
step 2 are untouched; they still route straight to `Terminal`.

```{r step3}
states_v3 <- states_v2

states_v3[["Tag Metformin Regimen"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "metformin", transition = direct("Treatment Review Delay")
)

states_v3[["Treatment Review Delay"]] <- create_state_settings("Delay",
  range = create_component_settings("range", low = 6, high = 18, unit = "months"),
  transition = direct("Metformin Switch Check")
)

states_v3[["Metformin Switch Check"]] <- create_state_settings("Simple",
  transition = create_transition_settings("distributed", options = list(
    list(transition = "Continue Metformin", distribution = 0.60),
    list(transition = "Switch To Sulfonylurea", distribution = 0.15),
    list(transition = "Switch To Third Agent", distribution = 0.15),
    list(transition = "Discontinue Metformin", distribution = 0.10)
  ))
)

states_v3[["Continue Metformin"]] <- create_state_settings("Simple", transition = direct("Terminal"))

states_v3[["Switch To Sulfonylurea"]] <- create_state_settings("MedicationEnd",
  medication_order = "Start Metformin", transition = direct("Start Switch Sulfonylurea")
)
states_v3[["Start Switch Sulfonylurea"]] <- create_state_settings("MedicationOrder",
  codes = list(sulfonylurea_code), reason = "Diabetes Onset", chronic = TRUE,
  transition = direct("Tag Switched Regimen (Sulfonylurea)")
)
states_v3[["Tag Switched Regimen (Sulfonylurea)"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "sulfonylurea (switched)", transition = direct("Terminal")
)

states_v3[["Switch To Third Agent"]] <- create_state_settings("MedicationEnd",
  medication_order = "Start Metformin", transition = direct("Start Third Agent")
)
states_v3[["Start Third Agent"]] <- create_state_settings("MedicationOrder",
  codes = list(third_agent_code), reason = "Diabetes Onset", chronic = TRUE,
  transition = direct("Tag Switched Regimen (Other)")
)
states_v3[["Tag Switched Regimen (Other)"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_regimen", value = "other", transition = direct("Terminal")
)

states_v3[["Discontinue Metformin"]] <- create_state_settings("MedicationEnd",
  medication_order = "Start Metformin", transition = direct("Terminal")
)

cat(build_module(name = "Cohort Building Vignette", states = states_v3))
```

A few things worth noticing:

- **`MedicationEnd.medication_order` names the state that started the drug being ended** --
  `"Start Metformin"`, the same state all three of these pathways came through -- not the drug
  itself. `validate_module()` checks that this name resolves to a real state the same way it
  checks any other transition target (see the explicit check below).
- The switched-to-sulfonylurea patients get their *own* `MedicationOrder` state
  (`"Start Switch Sulfonylurea"`), separate from the first-line `"Start Sulfonylurea"` in step 2,
  tagged with a distinguishable `diabetes_regimen` value (`"sulfonylurea (switched)"` vs.
  `"sulfonylurea"`). Nothing forces this split -- routing both pathways into the same
  `"Start Sulfonylurea"` state would work too, since Synthea resolves states by name and a state
  can have any number of incoming transitions. Keeping them separate here is a design choice: it
  lets an analysis distinguish "first-line sulfonylurea" from "switched to sulfonylurea after
  metformin" as two different subgroups, which is exactly the kind of thing a real cohort study
  cares about.
- **`"Discontinue Metformin"` is only a `MedicationEnd`** -- no new `MedicationOrder`, and,
  deliberately, no state re-tagging `diabetes_regimen` afterwards. It still reads `"metformin"`,
  the value `"Tag Metformin Regimen"` set on the way in. That's not an oversight; step 4 uses
  exactly this pathway to show why you shouldn't rely on that tag alone.

## Step 4: classify by the record, not by the path

`diabetes_regimen` records *which pathway a patient took*. That's useful, but it's not the same
thing as *what's actually true about their record right now* -- and step 3's `"Discontinue
Metformin"` pathway already made those two facts diverge: `diabetes_regimen` still says
`"metformin"` there, even though the medication itself was ended with nothing started in its
place.

So the last step classifies every patient by querying their actual medication state with
`ActiveMedication` logic, combined with `Or` (`vignette("logic-and-guards")`), rather than by
trusting the attribute set earlier:

```{r step4}
states_v4 <- states_v3

on_any_diabetes_medication <- create_logic_settings("Or", conditions = list(
  create_logic_settings("ActiveMedication", codes = list(metformin_code)),
  create_logic_settings("ActiveMedication", codes = list(sulfonylurea_code)),
  create_logic_settings("ActiveMedication", codes = list(third_agent_code))
))

states_v4[["Classify Treatment Status"]] <- create_state_settings("Simple",
  transition = create_transition_settings("conditional", options = list(
    list(condition = on_any_diabetes_medication, transition = "Treated Cohort"),
    list(transition = "Untreated Cohort") # condition omitted = the else branch
  ))
)
states_v4[["Treated Cohort"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_cohort", value = "treated", transition = direct("Terminal")
)
states_v4[["Untreated Cohort"]] <- create_state_settings("SetAttribute",
  attribute = "diabetes_cohort", value = "untreated", transition = direct("Terminal")
)

# Route every leaf that previously went straight to Terminal through classification first.
for (leaf in c(
  "Tag Sulfonylurea Regimen", "No Medication", "Continue Metformin",
  "Tag Switched Regimen (Sulfonylurea)", "Tag Switched Regimen (Other)",
  "Discontinue Metformin"
)) {
  states_v4[[leaf]]$direct_transition <- "Classify Treatment Status"
}

full_module <- build_module(name = "Cohort Building Vignette", states = states_v4)
cat(full_module)
```

Walk through what `diabetes_cohort` ends up as for each subgroup: `"No Medication"` never ordered
anything, so `on_any_diabetes_medication` is false -- `"untreated"`. `"Continue Metformin"` and
first-line `"Tag Sulfonylurea Regimen"` are each still on the drug they started -- `"treated"`.
Both switch pathways ended metformin and started a different drug -- still `"treated"`, correctly,
even though `diabetes_regimen` now reads `"sulfonylurea (switched)"` or `"other"` instead of
`"metformin"`. And `"Discontinue Metformin"` -- the pathway where `diabetes_regimen` is stale and
still says `"metformin"` -- correctly comes out `"untreated"` anyway, because `ActiveMedication`
checks the actual record, not the leftover tag. The classification is robust to exactly how a
patient got there, because it asks the record instead of the path.

## Checking the graph explicitly with `validate_module()`

`build_module()` already calls `validate_module()` internally (`validate = TRUE` by default), which
is why every `cat(build_module(...))` call above would have stopped with an error instead of
printing JSON if any of these six pathways had a dangling reference. You can also run that same
check yourself, directly, against a module list -- useful when you're assembling `states` across
several R objects or a script, and want to check the graph before deciding whether to render it to
JSON at all:

```{r validate-explicit}
full_module_list <- build_module(name = "Cohort Building Vignette", states = states_v4, as_json = FALSE)
isTRUE(validate_module(full_module_list))
```

`validate_module()` is what actually catches the mistake in a broken graph, not `build_module()`
itself -- `build_module()` just calls it and propagates the error. Reproducing the exact mistake
the first bullet in step 3 warned about, a typo'd `medication_order` reference, shows what that
looks like:

```{r validate-broken, error = TRUE}
broken_states <- states_v4
broken_states[["Discontinue Metformin"]]$medication_order <- "Start Metfromin" # typo
validate_module(list(states = broken_states))
```

## Sanity-checking cohort proportions before touching Synthea

`vignette("transition-reference")` notes that a `distributed_transition`'s `distribution` values
should sum to ~1.0 across each set of options. That constraint also means you can compute expected
cohort sizes by hand, straight from the numbers you just wrote, as a design-time check that the
population mix looks like what you intended -- without generating a single patient:

```{r sanity-check}
first_line <- c(metformin = 0.5, sulfonylurea = 0.3, none = 0.2)
metformin_outcome <- c(
  continues = 0.60, switches_sulfonylurea = 0.15,
  switches_other = 0.15, discontinues = 0.10
)

data.frame(
  final_regimen = c(
    "metformin", "sulfonylurea (first-line)", "sulfonylurea (switched)",
    "other (switched)", "discontinued (no replacement)", "none (first-line)"
  ),
  expected_share = c(
    first_line["metformin"] * metformin_outcome["continues"],
    first_line["sulfonylurea"],
    first_line["metformin"] * metformin_outcome["switches_sulfonylurea"],
    first_line["metformin"] * metformin_outcome["switches_other"],
    first_line["metformin"] * metformin_outcome["discontinues"],
    first_line["none"]
  ),
  row.names = NULL
)
```

That table should sum to 1 (it does), and `"none (first-line)"` staying exactly `0.2` -- unchanged
since step 2 -- is the arithmetic confirmation that adding the switching and discontinuation logic
in step 3 never touched the no-medication or first-line-sulfonylurea pathways at all. Two rows
("discontinued (no replacement)" and "none (first-line)") both end up `"untreated"` after step 4's
classification, even though only one of them was ever tagged that way along the path.

## Where to go from here

This module stops at classification, but nothing about the pattern is specific to diabetes or to
two drugs. The same three moves -- split into pathways with `distributed_transition`, tag each
pathway with `SetAttribute`, and later re-derive group membership from
`ActiveMedication`/`ActiveCondition` logic rather than trusting the tag alone -- extend to any
cohort with heterogeneous treatment pathways: a `Guard` (`vignette("logic-and-guards")`) could
hold patients on a drug until an `Observation` crosses some threshold before letting them switch;
a `conditional_transition` keyed on `PriorState` could route only patients who've actually been
through `"Switch To Third Agent"` into a separate follow-up pathway; a `distributed_transition`
whose probability comes from a `NamedDistribution` (`list(attribute = ..., default = ...)`; see
`vignette("transition-reference")`) could make the switch or discontinuation probability
itself depend on an attribute set earlier in the same module, instead of a fixed number; and
`validate_module()`, called explicitly as above, is worth running again after any such addition
before regenerating the JSON.
