---
title: "Component Reference: All Eight Value Shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Component Reference: All Eight Value Shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`vignette("layers")` introduces `create_component_settings()` and walks through three of its
eight shapes (`code`, `range`, `distribution`) in enough depth to show how the function works.
This vignette is the complete reference: one example for every shape, including the five
(`exact`, `date_input`, `sampled_data`, `attachment`, `io_mapper`) that don't otherwise appear
anywhere in this package's vignettes.

```{r setup}
library(syntheaModuleR)
```

## Quick reference

| `component`   | Required                  | Optional                                                                 | Array fields |
|---------------|---------------------------|----------------------------------------------------------------------------|---------------|
| `code`        | `system`, `code`, `display` |                                                                            |               |
| `range`       | `low`, `high`              | `unit`, `decimals`                                                        |               |
| `exact`       | `quantity`                 | `unit`                                                                    |               |
| `date_input`  | `year`, `month`, `day`     | `hour`, `minute`, `second`, `millisecond`                                 |               |
| `sampled_data`| `origin_value`, `attributes` | `factor`, `lower_limit`, `upper_limit`, `decimal_format`                | `attributes`  |
| `attachment`  |                            | `content_type`, `language`, `title`, `creation`, `height`, `width`, `frames`, `duration`, `pages`, and **exactly one of** `chart`, `url`, `data` | |
| `distribution`| `kind`                     | `round`, plus `kind`-specific `parameters` -- see `vignette("distribution")` | |
| `io_mapper`   | `type`                     | `from`, `to`, `from_list`, `from_exp`, `variance`, `vital_sign`            |               |

## `code` and `range` (recap)

Already introduced in `vignette("layers")` -- shown here just for completeness:

```{r code-range}
create_component_settings("code",
  system = "SNOMED-CT", code = "444316003", display = "Seasonal allergic rhinitis (disorder)"
)

create_component_settings("range", low = 3, high = 7, unit = "days")
```

## `exact`

The single-value counterpart to `range` -- one `quantity`, with an optional `unit`. It's the
shape a `Procedure`'s `duration` or a `Death`'s `exact` expects:

```{r exact}
create_component_settings("exact", quantity = 30, unit = "minutes")
```

## `date_input`

A calendar date/time, used where a `Date` Logic condition needs a literal point in time to
compare against rather than an offset from the simulated "now". `year`, `month`, and `day` are
required; the time-of-day fields are all optional and default to midnight if omitted:

```{r date_input}
create_component_settings("date_input", year = 2021, month = 3, day = 15, hour = 8, minute = 30)
```

## `sampled_data`

Backs an `Observation`'s `sampled_data` field -- a time series of measurements (an HL7
`SampledData`-shaped value) rather than a single point value, tagged with the attribute name(s)
each sampled column corresponds to:

```{r sampled_data}
create_component_settings("sampled_data",
  origin_value = 0, attributes = list("ecg_lead_ii"),
  factor = 0.01, lower_limit = -200, upper_limit = 200, decimal_format = "0.00"
)
```

## `attachment`

An embedded chart, image, or file reference -- used inside an `Observation`'s `attachment`
field. No field is individually required, but exactly one of `chart`, `url`, or `data` must be
given; the Java engine's own `Attachment.validate()` rejects both zero and more than one, same as
this package's `one_of` check:

```{r attachment}
create_component_settings("attachment",
  url = "https://example.org/growth-chart.png", content_type = "image/png", title = "Growth Chart"
)
```

```{r attachment-error, error = TRUE}
create_component_settings("attachment",
  chart = "base64-encoded-chart-data", url = "https://example.org/growth-chart.png"
)
```

## `io_mapper`

Only meaningful inside a `Physiology` state's `inputs`/`outputs` (see `vignette("layers")`'s note
that `Physiology` is config-gated and out of scope for most modules). `type` selects an `IoType`
(e.g. `"ATTRIBUTE"`, `"VITAL_SIGN"`) -- this package doesn't validate which strings are legal
there, since only `IoMapper`'s core fields are modeled, not its nested `PreGenerator` type:

```{r io_mapper}
create_component_settings("io_mapper", type = "VITAL_SIGN", vital_sign = "Heart Rate", to = "heart_rate")
```

## `distribution`

The one irregular shape -- exact-case `kind` values, a `parameters` map that isn't
snake_cased -- gets a full vignette of its own: `vignette("distribution")`.

## Where to go from here

`vignette("layers")` shows how `code` and `range` nest inside real states
(`ConditionOnset.codes`, `Delay.range`); `vignette("distribution")` does the same for
`distribution`. The remaining shapes here -- `exact`, `date_input`, `sampled_data`, `attachment`,
`io_mapper` -- are less commonly needed, so nesting them into a full worked module is left as an
exercise: try building an `Observation` state with an `attachment`, or a `Procedure` with an
`exact` duration, using `create_state_settings()` the same way `vignette("layers")` does.
