---
title: "The Distribution Component: Modeling Randomness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Distribution Component: Modeling Randomness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")`. It goes deep on one component --
`distribution` -- which `vignette("layers")` flagged as the one irregular shape in the whole
toolkit, and worth a closer look on its own.

```{r setup}
library(syntheaModuleR)
```

## Why it's irregular

Every other component (`code`, `range`, `exact`, ...) is a flat object whose field names are
declared Java fields, so Gson's `LOWER_CASE_WITH_UNDERSCORES` policy rewrites them to
snake_case -- which is exactly what this toolkit's field names already are everywhere else.

`Distribution` (`Distribution.java`) breaks that pattern in two ways at once:

1. Its `kind` field (`Distribution.Kind`, a Java enum) is serialized by `.name()`, not rewritten
   by the naming policy -- so JSON values stay **exact-case upper**: `"GAUSSIAN"`, not
   `"gaussian"` or `"Gaussian"`.
2. Everything else it needs lives in a `parameters` map -- a `HashMap<String, Double>`, not a set
   of declared fields -- so the naming policy never touches those keys either. `mean` and `low`
   happen to already be lowercase, but `standardDeviation` stays camelCase.

`create_component_settings("distribution", kind = ..., ...)` handles this with its own code path,
separate from the generic schema-driven assembly every other component uses: `kind` selects a
`.distribution_param_schema` entry, and every other argument you pass becomes a `parameters` map
entry, validated against that entry's required/optional keys.

## The five kinds

| `kind`        | Required `parameters`   | Optional `parameters` |
|---------------|--------------------------|------------------------|
| `EXACT`       | `value`                  |                        |
| `UNIFORM`     | `low`, `high`            |                        |
| `GAUSSIAN`    | `mean`, `standardDeviation` | `min`, `max`        |
| `EXPONENTIAL` | `mean`                   |                        |
| `TRIANGULAR`  | `min`, `mode`, `max`     |                        |

```{r exact}
create_component_settings("distribution", kind = "EXACT", value = 5)
```

```{r uniform}
create_component_settings("distribution", kind = "UNIFORM", low = 10, high = 20)
```

```{r gaussian}
# min/max here clamp the sample to a range -- a truncated Gaussian, not a hard requirement.
create_component_settings("distribution",
  kind = "GAUSSIAN",
  mean = 120, standardDeviation = 15, min = 90, max = 180
)
```

```{r exponential}
create_component_settings("distribution", kind = "EXPONENTIAL", mean = 30)
```

```{r triangular}
create_component_settings("distribution", kind = "TRIANGULAR", min = 14, mode = 21, max = 28)
```

An optional `round` flag (a plain boolean, not a `parameters` entry) rounds the sampled value to
the nearest integer -- useful when the value feeds something that shouldn't come out fractional,
like a dose count:

```{r round}
create_component_settings("distribution", kind = "UNIFORM", low = 1, high = 4, round = TRUE)
```

## Two gotchas, live

`kind` must match one of the five names above exactly -- lower- or mixed-case is rejected before
it ever reaches JSON, since it wouldn't round-trip through the Java enum:

```{r kind-case-error, error = TRUE}
create_component_settings("distribution", kind = "gaussian", mean = 0, standardDeviation = 1)
```

And because `standardDeviation` is a real R argument name on `create_component_settings()` (not
snake_case, unlike its neighbors), typing the snake_case version you'd reflexively type for every
other field in this package doesn't get caught by this toolkit's own validation at all -- it's not
a recognized argument, so R itself refuses the call:

```{r snake-case-error, error = TRUE}
create_component_settings("distribution", kind = "GAUSSIAN", mean = 0, standard_deviation = 1)
```

## Where `distribution` nests

`distribution` is one option among several on a handful of states, and each state constrains it
differently:

| State       | Field's `one_of` group                              | Enforced by this package? |
|-------------|-------------------------------------------------------|----------------------------|
| `Delay`     | exactly one of `range`, `exact`, `distribution`        | Yes                        |
| `VitalSign` | exactly one of `exact`, `range`, `expression`, `distribution` | Yes                |
| `Procedure` | at most one of `duration`, `distribution`              | Yes                        |
| `SetAttribute` | one value source among several (see precedence below) | No -- Java-engine runtime behavior only |
| `Symptom`   | `exact`, `range`, `distribution` all just optional     | No constraint              |

`Delay` and `VitalSign` are enforced with the same `one_of` machinery as every other constraint in
this package -- pass two of them and `create_state_settings()` rejects the call before it ever
reaches `build_module()`:

```{r delay-oneof-error, error = TRUE}
create_state_settings("Delay",
  exact = create_component_settings("exact", quantity = 5, unit = "days"),
  distribution = create_component_settings("distribution", kind = "EXACT", value = 5),
  transition = create_transition_settings("direct", to = "Next")
)
```

`SetAttribute` is the one exception worth calling out explicitly: it accepts `distribution`
alongside `value`, `value_code`, `value_attribute`, `range`, `expression`, and `series_data`, but
this package does not enforce a `one_of` on them, because the Java engine doesn't treat it as a
validation error either -- it resolves value sources by precedence instead (first match wins):
`expression` > `range` > `series_data` > `distribution` > `value_code` > `value_attribute` >
literal `value`. Passing both `range` and `distribution` to a `SetAttribute` is valid, syntactically;
`range` will just always win, silently.

## Worked example: a two-dose delay and a sampled vital sign

```{r worked-example}
dose_interval <- create_component_settings("distribution",
  kind = "TRIANGULAR",
  min = 17, mode = 21, max = 28
) # realistic real-world dosing window, peaked at the trial's 21 days

bmi_sample <- create_component_settings("distribution",
  kind = "GAUSSIAN",
  mean = 27, standardDeviation = 5, min = 15, max = 50
)

states <- list(
  Initial = create_state_settings("Initial",
    transition = create_transition_settings("direct", to = "Record BMI")
  ),
  "Record BMI" = create_state_settings("VitalSign",
    vital_sign = "BMI", unit = "kg/m2", distribution = bmi_sample,
    transition = create_transition_settings("direct", to = "Dose Interval")
  ),
  "Dose Interval" = create_state_settings("Delay",
    distribution = dose_interval, unit = "days",
    transition = create_transition_settings("direct", to = "Terminal")
  ),
  Terminal = create_state_settings("Terminal")
)

module_json <- build_module(name = "Distribution Vignette Example", states = states)
cat(module_json)
```

Both states validated against their own `one_of` constraint (each has exactly one value source),
and `build_module()` validated the graph on top of that -- the same two-level check
(`create_state_settings()` per-state, `validate_module()` graph-wide) every module in this package
goes through.

## Where to go from here

`SetAttribute`'s full precedence chain (`expression` > `range` > `series_data` > `distribution` >
`value_code` > `value_attribute` > `value`) deserves its own worked example, as does a
side-by-side of `build_disease_module()`'s onset -> encounter -> resolve/death recipe against
hand-composing the same shape with `build_module()` directly.
