---
title: "The Frontend Layer: From One Condition to a Full Protocol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Frontend Layer: From One Condition to a Full Protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`vignette("layers")` and `vignette("cohort-building")` build modules with the Engine layer
directly: every state gets its own `create_state_settings()` call, and every transition target is
a hand-typed state name. That's the right level when you need exact control, but it gets verbose
fast once a module has more than a handful of states.

The **Frontend layer** (`blocks.R`, `combinators.R`) sits on top of the Engine layer and adds
nothing to the JSON grammar itself -- every function here still only ever emits
`create_state_settings()` output under the hood. What it adds is a shared contract, a **fragment**:

```r
list(states, entry, exit)
```

a named list of Engine-layer states, the name of the state to route *into* (`entry`), and the name
of the state whose transition is still unresolved (`exit`) -- left pointing at the literal
placeholder `"__PENDING__"` until a combinator or `build_cohort_module()` patches it. Every
`create_*()` leaf in `blocks.R` returns one; every combinator in `combinators.R` consumes one or
more and returns a new one. Because the contract never changes, fragments nest arbitrarily deep:
the output of `pathways()` is exactly as valid an input to `chain()` or another `pathways()` as a
single `create_condition()` call is.

This vignette builds up from the smallest possible fragment to a full protocol-shaped cohort
module, using worked examples from `experimentation/notes/protocol-intake-checklist.md` -- a
checklist for turning a real study protocol into a cohort module, distilled from three actual
protocols kept in this repository (`experimentation/covid-19.docx`,
`experimentation/immunocompromised.docx`, `experimentation/valproate.docx`). As that checklist
warns, every medical code used below is an illustrative placeholder, not verified against a real
codeset -- the same disclaimer `cohort-building.Rmd` and `example_trial_emulation.R` carry for
their own example codes.

```{r setup}
library(syntheaModuleR)
```

## Level 1: the smallest fragment -- one condition

`create_condition()` is a leaf: onset, diagnosis encounter, optional resolution, wrapped as one
fragment. `build_cohort_module()` is the Frontend layer's finisher -- the only place `Initial`/
`Terminal` get invented -- so a single leaf is already a complete module:

```{r level1}
diabetes_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "44054006", display = "Diabetes mellitus type 2 (disorder)")

cohort_v1 <- create_condition("Diabetes", diabetes_code, diagnosis = "wellness")
cat(build_cohort_module("Diabetes Cohort", cohort_v1))
```

Compare this to `vignette("cohort-building")`'s step 1: same four clinical states (onset,
encounter, encounter end, terminal), but here they came from one function call instead of four
`create_state_settings()` calls plus manually wiring `Initial`/`Terminal`.

## Level 2: `chain()`/`add()` -- sequencing two fragments

`chain()` wires each fragment's `exit` to the next fragment's `entry`; `add()` is the same thing,
spelled for piping onto something you already have. Adding a chronic medication after the
diagnosis is one more leaf and one `add()`:

```{r level2}
metformin_code <- create_component_settings("code",
  system = "RxNorm", code = "860975", display = "Metformin hydrochloride 500 MG Oral Tablet")

cohort_v2 <- create_condition("Diabetes", diabetes_code, diagnosis = "wellness") |>
  add(create_medication("Metformin", metformin_code, condition = "Diabetes", duration = "long"))

cat(build_cohort_module("Diabetes Cohort", cohort_v2))
```

`create_medication()`'s `condition = "Diabetes"` resolves as the `MedicationOrder`'s `reason` --
it's the same string as `create_condition()`'s own `label` above, not a coincidence: the onset
state's name *is* the handle other blocks reference.

## Level 3: `pathways()` -- probability-weighted exposure arms

Real cohorts are rarely one drug for everyone. `pathways()` takes a named list of fragments (an
entry can be `NULL` for "nothing happens on this arm"), a matching `shares` vector, and an optional
`attribute` that auto-tags which arm a patient went through -- one `distributed_transition` "choice"
state fanning out, one "join" state everything rejoins into:

```{r level3}
sulfonylurea_code <- create_component_settings("code",
  system = "RxNorm", code = "310488", display = "Glipizide 5 MG Oral Tablet")

first_line <- pathways("First-Line Treatment",
  options = list(
    metformin    = create_medication("Metformin", metformin_code, condition = "Diabetes", duration = "long"),
    sulfonylurea = create_medication("Sulfonylurea", sulfonylurea_code, condition = "Diabetes", duration = "long"),
    none         = NULL
  ),
  shares = c(metformin = 0.5, sulfonylurea = 0.3, none = 0.2),
  attribute = "diabetes_regimen"
)

cohort_v3 <- chain(create_condition("Diabetes", diabetes_code, diagnosis = "wellness"), first_line)
cat(build_cohort_module("Diabetes Cohort", cohort_v3))
```

As in `vignette("cohort-building")`'s step 2, `"none"` isn't a flag set afterward -- it's an arm
that never issues a `MedicationOrder` at all, expressed here as a bare `NULL` instead of a fourth
state.

## Level 4: switching mid-pathway -- fragments nest

`vignette("cohort-building")`'s step 3 redirects a state by hand to add a switching sub-graph
partway through. With the Frontend layer, that sub-graph is just another fragment, and "partway
through" is just `add()`ing it onto the metformin arm *before* that arm is handed to `pathways()`:

```{r level4}
third_agent_code <- create_component_settings("code",
  system = "RxNorm", code = "1545653", display = "Empagliflozin 25 MG Oral Tablet")

metformin_arm <- create_medication("Metformin", metformin_code, condition = "Diabetes", duration = "long") |>
  add(create_delay(6, 18, "months", label = "Treatment Review Delay")) |>
  add(pathways("Metformin Switch",
    options = list(
      continues              = NULL,
      switches_sulfonylurea  = create_discontinue("Discontinue Metformin For Sulfonylurea", medication = "Metformin") |>
        add(create_medication("Switch To Sulfonylurea", sulfonylurea_code, condition = "Diabetes", duration = "long")),
      switches_other         = create_discontinue("Discontinue Metformin For Third Agent", medication = "Metformin") |>
        add(create_medication("Switch To Third Agent", third_agent_code, condition = "Diabetes", duration = "long")),
      discontinues           = create_discontinue("Discontinue Metformin", medication = "Metformin")
    ),
    shares = c(continues = 0.60, switches_sulfonylurea = 0.15, switches_other = 0.15, discontinues = 0.10)
  ))

first_line_v4 <- pathways("First-Line Treatment",
  options = list(metformin = metformin_arm,
                 sulfonylurea = create_medication("Sulfonylurea", sulfonylurea_code, condition = "Diabetes", duration = "long"),
                 none = NULL),
  shares = c(metformin = 0.5, sulfonylurea = 0.3, none = 0.2),
  attribute = "diabetes_regimen")

cohort_v4 <- chain(create_condition("Diabetes", diabetes_code, diagnosis = "wellness"), first_line_v4)
```

Nothing about `pathways()` or `chain()` changed between level 3 and level 4 -- only the fragment
passed in as the `metformin` option got more elaborate. That's the point of the fragment contract:
a combinator never needs to know whether an option is a single `create_medication()` call or an
entire nested `pathways()` sub-graph three levels deep.

## Level 5: `classify()` -- classify by record, not by path

`diabetes_regimen` (set by `pathways()`'s `attribute =`) records *which arm a patient took*, not
*what's true about their medication record right now* -- `vignette("cohort-building")`'s step 4
makes exactly this point: the discontinued-metformin pathway still carries the stale tag
`"metformin"`. `classify()` is `pathways()`'s condition-driven twin: instead of a probability split,
it routes on a `Logic` condition, with the **first** named option taken when the condition is true
and the **second** the `else` fallback.

```{r level5}
on_any_diabetes_med <- create_logic_settings("Or", conditions = list(
  create_logic_settings("ActiveMedication", codes = list(metformin_code)),
  create_logic_settings("ActiveMedication", codes = list(sulfonylurea_code)),
  create_logic_settings("ActiveMedication", codes = list(third_agent_code))
))

treatment_status <- classify("Treatment Status", condition = on_any_diabetes_med,
  options = list(
    treated   = create_tag("Treated", attribute = "diabetes_cohort", value = "treated"),
    untreated = create_tag("Untreated", attribute = "diabetes_cohort", value = "untreated")
  ))

cohort_v5 <- chain(create_condition("Diabetes", diabetes_code, diagnosis = "wellness"),
                    first_line_v4, treatment_status)
final_module <- build_cohort_module("Diabetes Cohort", cohort_v5, as_json = FALSE)
isTRUE(validate_module(final_module))
```

Five combinator calls (`create_condition()`, `pathways()` nested inside `pathways()`, `classify()`,
tied together with `chain()`) reproduce the entire growing example `vignette("cohort-building")`
builds with eighteen individual `create_state_settings()` calls and several manual re-pointed
`direct_transition`s -- same JSON shape, same validated module, far less to type or get wrong by
hand.

## Level 6: population eligibility, lifted from a real protocol

Every example so far skipped *who gets into the cohort at all*. That's `create_population()` (a
flat demographic `AND`) and `create_guard()` (anything else -- a condition that must pre-exist, a
washout window, a composite `Or()`/`And()`/`Not()` tree) -- both documented in
`experimentation/notes/protocol-intake-checklist.md`'s section 1, "Population eligibility."

The checklist's valproate worked example (`experimentation/valproate.docx`: paternal valproate
exposure and offspring outcomes) is reproduced here verbatim. It's chosen over the checklist's
COVID-19 and immunocompromised examples deliberately: in *this* protocol, "contraindication" and
"infertility" really are single, atomic, code-list-driven concepts, not composite algorithms the
way "immunocompromised" turned out to be in the other two (see the checklist's "Dig deeper" note
if you're adapting your own protocol and a criterion reads as one adjective in the summary text):

```{r level6}
epilepsy_code <- create_component_settings("code", system = "SNOMED-CT", code = "84757009",
  display = "Generalised epilepsy (disorder)")
valproate_allergy <- create_component_settings("code", system = "SNOMED-CT", code = "428607008",
  display = "Allergy to valproic acid (illustrative -- one of a longer contraindication codelist)")
hepatic_contraindication <- create_component_settings("code", system = "SNOMED-CT", code = "197321007",
  display = "Acute hepatic failure (disorder) -- illustrative; valproate is contraindicated in hepatic impairment")
male_infertility <- create_component_settings("code", system = "SNOMED-CT", code = "237186005",
  display = "Male infertility (disorder)")
vasectomy <- create_component_settings("code", system = "SNOMED-CT", code = "16621005",
  display = "Vasectomy (procedure) -- illustrative; a permanent-infertility procedure, not just a diagnosis")

valproate_population <- chain(
  create_population("Demographic Eligibility",
                     age = list(operator = ">", quantity = 18, unit = "years"), gender = "M"),
  create_guard(
    create_logic_settings("ActiveCondition", codes = list(epilepsy_code)),
    label = "Epilepsy Diagnosis Present"),
  create_guard(
    create_logic_settings("Not", condition =
      create_logic_settings("ActiveCondition", codes = list(valproate_allergy, hepatic_contraindication))),
    label = "Exclude Valproate-Levetiracetam Contraindication"),
  create_guard(
    create_logic_settings("Not", condition =
      create_logic_settings("Or", conditions = list(
        create_logic_settings("ActiveCondition", codes = list(male_infertility)),
        create_logic_settings("ActiveCondition", codes = list(vasectomy))
      ))),
    label = "Exclude Infertility")
)
```

`chain()` on four fragments here does the same job it did sequencing clinical states earlier --
`create_population()` returns `NULL` if every filter is omitted (a true no-op `chain()` silently
skips), and each `create_guard()` is a `Guard` state that re-tests its condition every timestep
until it passes, not a one-shot branch. This fragment isn't a complete module by itself (nothing
ends its `exit`'s pending transition yet) -- it's a population gate meant to be `chain()`ed in
front of whatever the rest of the module does, exactly like every clinical fragment above.

## Level 7: putting it together -- a full protocol-shaped module

The checklist's pipeline runs population (section 1) -> condition -> exposure (section 4) ->
switching (section 5) -> outcome/follow-up (sections 6-7) -> final classification (section 8). The
valproate protocol itself supplies two more shapes worth showing that none of the levels above
needed yet:

- **An active-comparator exposure**, not an untreated arm -- the protocol uses levetiracetam
  "since epileptic patients are unlikely left untreated." That's `pathways()` with two medication
  arms and no `NULL` option, unlike the diabetes example's `"none"` arm above.
- **Gap-tolerant exposure accumulation** -- the checklist flags "a refill gap of up to N days
  doesn't count as discontinuation" as the one shape needing `create_counter()` +
  `repeat_until()` rather than a plain `create_medication()`. `repeat_until()` is the one
  combinator that wires a transition *backward*: it runs a `body` fragment once, checks an `until`
  condition, and only exits the loop once that condition is satisfied.

```{r level7}
valproate_code <- create_component_settings("code", system = "RxNorm", code = "9825",
  display = "Valproic acid (illustrative)")
levetiracetam_code <- create_component_settings("code", system = "RxNorm", code = "114477",
  display = "Levetiracetam (illustrative active comparator)")
hepatic_outcome_code <- create_component_settings("code", system = "SNOMED-CT", code = "235856003",
  display = "Hepatic impairment (disorder) -- illustrative safety outcome")

# Gap-tolerant persistence: check in every 30 days: reaching 6 qualifying checks (~6 months of
# refills without a disqualifying gap) is this illustrative stand-in for "still persistent."
persistence_loop <- repeat_until(
  body = chain(
    create_delay(30, 30, "days", label = "Refill Window"),
    create_counter("valproate_refill_checks", action = "increment", label = "Count Refill Check")
  ),
  until = create_logic_settings("Attribute", attribute = "valproate_refill_checks",
                                 operator = ">=", value = 6),
  label = "Valproate Persistence")

exposure_arm <- pathways("Exposure Arm",
  options = list(
    valproate    = create_medication("Valproate", valproate_code,
                     condition = "Epilepsy Diagnosis Present", duration = "long") |>
      add(persistence_loop),
    levetiracetam = create_medication("Levetiracetam", levetiracetam_code,
                     condition = "Epilepsy Diagnosis Present", duration = "long")
  ),
  shares = c(valproate = 0.5, levetiracetam = 0.5),
  attribute = "exposure_arm")

follow_up_cap <- create_delay(90, 90, "days", label = "Administrative Follow-Up Cap")

outcome_status <- classify("Outcome Status",
  condition = create_logic_settings("ActiveCondition", codes = list(hepatic_outcome_code)),
  options = list(
    outcome    = create_tag("Outcome Occurred", attribute = "outcome_status", value = "event"),
    no_outcome = create_tag("No Outcome", attribute = "outcome_status", value = "censored")
  ))

protocol_module <- chain(valproate_population, exposure_arm, follow_up_cap, outcome_status)
final <- build_cohort_module("Valproate vs Levetiracetam in Epilepsy", protocol_module, as_json = FALSE)
isTRUE(validate_module(final))
length(final$states)
```

Note `condition = "Epilepsy Diagnosis Present"` on both `create_medication()` calls: it resolves
against the `Guard` state of that name from `valproate_population` -- `MedicationOrder.reason`
resolution falls back to "a `ConditionOnset` state of that name," but here epilepsy is a
*pre-existing* condition being checked, not one this module diagnoses, so there's no
`create_condition()` state named `"Epilepsy Diagnosis Present"` for it to resolve to cleanly; the
Engine's own resolution order (`vignette("state-reference")`) also tries a plain person attribute
first, which this population `Guard`'s label happens not to set -- worth flagging deliberately
(the same way the checklist's own worked examples flag every simplification) since it's the kind
of subtlety `validate_module()`'s state-name-reference check catches for *transition* targets but
has no way to catch for a `reason` string, which is free text checked at generation time by
Synthea, not at module-build time by this toolkit.

## Where to go from here

Every level here used a fragment straight out of `blocks.R`, composed with one of the four
combinators. Nothing in `pathways()`/`classify()`/`repeat_until()`/`chain()` cares how deep or
shallow the fragments underneath them are -- level 4's switching sub-graph inside a `pathways()`
option and level 7's `repeat_until()` loop inside a *different* `pathways()` option compose the
same way a single leaf does. `experimentation/notes/protocol-intake-checklist.md` walks the full
extraction process section by section (including the COVID-19 and immunocompromised protocols'
own population-eligibility examples, and the "Extract" / "Decide" / "Dig deeper" / "Out of scope"
distinctions worth knowing before turning your own protocol into a module) -- start there when
adapting a real study protocol instead of the illustrative diabetes/valproate examples here.
