---
title: "Package Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package builds valid [Synthea](https://github.com/synthetichealth/synthea) Generic Module
Framework (GMF) module JSON files from R. It does not run Synthea -- it only produces the JSON
file; generating patients from it needs a real Synthea checkout (`./run_synthea -m <name>`). Every
field name and constraint it knows about was extracted directly from Synthea's Java engine
(`org.mitre.synthea.engine`), not guessed from example modules.

Five small functions, one per layer, cover the whole grammar:

```
components  ->  logic  ->  transitions  ->  states  ->  module
```

That arrow shows *nesting* order -- each thing on the left can appear inside the thing to its
right. It's not the order you'll actually work in: `states` is what you write the most, since a
named list of states is what you hand to `build_module()` directly; components, logic, and
transitions only exist to fill in a state's fields. `vignette("layers")` introduces the five
layers in that states-first order, not bottom-up.

```{r setup}
library(syntheaModuleR)
```

The smallest possible valid module is just the two states every module needs -- `Initial` and one
`Terminal` -- wired directly together:

```{r minimal}
minimal_module <- build_module(
  name = "Minimal Example",
  states = list(
    Initial = create_state_settings("Initial",
      transition = create_transition_settings("direct", to = "Terminal")),
    Terminal = create_state_settings("Terminal")
  )
)
cat(minimal_module)
```

Everything else this package can do is variations and elaborations on that same pattern: more
states, richer transitions, `Logic` conditions gating them, `Component` value shapes filling them
in. The vignettes below build that up states-first, the same way you'd actually write a module.

## Read these in order

1. **[The Five Layers: Components, Logic, Transitions, States, Modules](layers.html)** -- start
   here. Starts with `States` -- the layer you'll actually spend the most time writing -- then
   unpacks what nests inside a state (`Components`, `Logic`, `Transitions`), before assembling
   everything into a `Module` at the end.
2. **[State Reference: All 31 State Types](state-reference.html)** -- the complete reference for
   `create_state_settings()`, including the states -- `CallSubmodule`, `Physiology`, `Counter`,
   `AllergyOnset`, `CarePlanStart`, `Vaccine`, and more -- `vignette("layers")` doesn't cover.
3. **[SetAttribute: Value Source Precedence](set-attribute-precedence.html)** -- the one state
   whose multiple value-source fields (`value`, `range`, `expression`, `distribution`, ...) are
   **not** mutually exclusive at build time, and the precedence order the Java engine uses to
   pick a winner when more than one is given.
4. **[Component Reference: All Eight Value Shapes](component-reference.html)** -- the complete
   reference for `create_component_settings()`'s eight shapes, including the five (`exact`,
   `date_input`, `sampled_data`, `attachment`, `io_mapper`) `vignette("layers")` doesn't cover.
5. **[The Distribution Component: Modeling Randomness](distribution.html)** -- the one
   irregular component shape (`EXACT`/`UNIFORM`/`GAUSSIAN`/`EXPONENTIAL`/`TRIANGULAR`), and where
   it nests inside `Delay`, `VitalSign`, and `Procedure`.
6. **[Transition Reference: All Six Kinds](transition-reference.html)** -- the complete
   reference for `create_transition_settings()`, including `complex`, `lookup_table`, and
   `type_of_care`, which `vignette("layers")` only describes.
7. **[Logic Reference: All 21 Predicate Types](logic-reference.html)** -- the complete reference
   for `create_logic_settings()`'s leaf predicates -- `Gender`, `Date`, `PriorState`,
   `ActiveAllergy`, and more -- that `vignette("logic-and-guards")` doesn't cover.
8. **[Logic Combinators and Guards](logic-and-guards.html)** -- the predicate sublanguage in
   depth: `And`/`Or`/`Not`/`AtLeast`/`AtMost`, and the `Guard` state that re-tests a condition
   every timestep rather than branching once.
9. **[build_disease_module() vs. build_module(): Recipe vs. Hand-Composed](recipe-vs-compose.html)**
   -- the onset -> encounter -> resolve/death convenience recipe, proven byte-for-byte identical
   to hand-composing the same shape, and exactly where its coverage ends.
10. **[Building Cohorts Incrementally: Diagnosis, Treatment Pathways, and Switching](cohort-building.html)**
    -- a worked, growing example: a diagnosis, a split into two first-line drugs (plus a
    no-medication pathway), a subset that switches therapy or discontinues it, and a final
    classification that reads the medication record instead of trusting the pathway a patient
    took -- plus an explicit `validate_module()` check.
11. **[The Frontend Layer: From One Condition to a Full Protocol](frontend-layer.html)** -- the
    same cohort-building shape as above, but composed from `blocks.R`/`combinators.R` fragments
    instead of individual `create_state_settings()` calls, growing from a single
    `create_condition()` up to a full protocol-shaped module (population eligibility, an
    active-comparator exposure, gap-tolerant persistence via `repeat_until()`, and a final
    `classify()`) lifted from the real study protocols in `experimentation/`.

## Quick reference

| Layer        | Function                          | Covered in                                    |
|--------------|------------------------------------|------------------------------------------------|
| States       | `create_state_settings()`          | [layers](layers.html), [state-reference](state-reference.html), [set-attribute-precedence](set-attribute-precedence.html) |
| Components   | `create_component_settings()`      | [layers](layers.html), [component-reference](component-reference.html), [distribution](distribution.html) |
| Logic        | `create_logic_settings()`          | [layers](layers.html), [logic-reference](logic-reference.html), [logic-and-guards](logic-and-guards.html) |
| Transitions  | `create_transition_settings()`     | [layers](layers.html), [transition-reference](transition-reference.html) |
| Module       | `build_module()`, `validate_module()`, `write_module_json()` | [layers](layers.html), [recipe-vs-compose](recipe-vs-compose.html) |
| Recipe       | `build_disease_module()`           | [recipe-vs-compose](recipe-vs-compose.html)    |

## Package layout

```
R/
  components.R   .component_schema + .distribution_param_schema + create_component_settings()
  logic.R        .logic_schema + create_logic_settings()
  transitions.R  .transition_schema + create_transition_settings()
  states.R       .state_schema + create_state_settings()
  module.R       build_module(), validate_module(), write_module_json(), build_disease_module()

build_disease_module.R      loader -- sources R/*.R in dependency order
example_generate.R          uses build_disease_module() (the recipe)
example_compose.R           uses build_module() + create_state_settings() directly
example_trial_emulation.R   a 47-state worked example (target-trial-emulation design)
```

Every example script -- and every vignette in this package -- sources
`build_disease_module.R`, never the individual `R/*.R` layer files directly.
