---
title: "Logic Combinators and Guards"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logic Combinators and Guards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")`. It goes deeper on one layer -- Logic --
and on the one state type that exists specifically to consume it in a loop: `Guard`.

```{r setup}
library(syntheaModuleR)
```

## Leaf predicates, briefly

`create_logic_settings(condition_type, ...)` builds a single Logic condition. Most of its 21
`condition_type`s are leaves that test one fact about the simulated person directly:

| `condition_type`                                          | Tests                                                    |
|-------------------------------------------------------------|-----------------------------------------------------------|
| `Gender`, `Age`, `Date`, `Race`, `SocioeconomicStatus`       | Demographics / simulated calendar date                    |
| `Attribute`, `VitalSign`, `Symptom`                          | A person attribute / vital sign / symptom severity value  |
| `Observation`                                                | A prior Observation state's recorded value                |
| `ActiveCondition`, `ActiveAllergy`, `ActiveMedication`, `ActiveCarePlan` | Whether something is currently active on the record |
| `PriorState`                                                  | Whether a named state was already visited (`since`/`within`) |
| `True`, `False`                                               | Constants -- for testing, not expected in a real module    |

```{r leaf-examples}
adult <- create_logic_settings("Age", operator = ">=", quantity = 18, unit = "years")
female <- create_logic_settings("Gender", gender = "F")
has_diabetes <- create_logic_settings("ActiveCondition",
  codes = list(create_component_settings("code",
    system = "SNOMED-CT", code = "44054006", display = "Diabetes mellitus type 2 (disorder)"
  ))
)
```

Four of these leaves (`Observation`, `ActiveCondition`, `ActiveAllergy`, `ActiveMedication`,
`ActiveCarePlan`) share the same constraint: **exactly one** of `codes` or
`referenced_by_attribute` -- never both, never neither. `create_logic_settings()` enforces this
the same way every other `one_of` group is enforced, as a build-time error:

```{r observation-oneof-error, error = TRUE}
create_logic_settings("Observation", operator = "==", value = 5)
```

`Observation` has a sharper gotcha the schema can't catch: the Java engine throws a
`NullPointerException` -- not a clean validation failure -- if the referenced Observation state
hasn't actually run yet by the time this condition is evaluated, unless `operator` is
`"is nil"`/`"is not nil"`. Guard a downstream `Observation` check with `And(PriorState(...), ...)`
(see below) rather than relying on ordering alone.

## The combinators: building a tree

Five `condition_type`s don't test anything themselves -- they combine other Logic results into a
tree. This is the actual point of this vignette: composition.

**`And` / `Or`** take a list of nested conditions via `conditions = list(...)`, all of which (`And`)
or any one of which (`Or`) must hold:

```{r and-or}
adult_female <- create_logic_settings("And", conditions = list(adult, female))

high_risk <- create_logic_settings("Or", conditions = list(
  has_diabetes,
  create_logic_settings("Age", operator = ">=", quantity = 65, unit = "years")
))
```

**`Not`** takes exactly one nested condition, singular -- `condition =`, not `conditions =`:

```{r not}
not_diabetic <- create_logic_settings("Not", condition = has_diabetes)
```

**`AtLeast` / `AtMost`** generalize `And`/`Or` to a quorum: `conditions` plus a `minimum` (for
`AtLeast`) or `maximum` (for `AtMost`) count of how many of those nested conditions must currently
evaluate to true. `And` is `AtLeast` with `minimum` equal to the full count; `Or` is `AtLeast` with
`minimum = 1`.

```{r atleast}
two_of_three_risk_factors <- create_logic_settings("AtLeast", minimum = 2, conditions = list(
  has_diabetes,
  create_logic_settings("Age", operator = ">=", quantity = 65, unit = "years"),
  create_logic_settings("VitalSign", vital_sign = "BMI", operator = ">=", value = 30)
))
```

Because every combinator argument is itself a `create_logic_settings()` result, these nest to
arbitrary depth -- an `Or` of `And`s of `Not`s composes exactly the way you'd build the equivalent
boolean expression in R itself:

```{r nested-tree}
eligible <- create_logic_settings("And", conditions = list(
  adult,
  create_logic_settings("Not", condition = create_logic_settings("Or", conditions = list(
    has_diabetes,
    create_logic_settings("ActiveAllergy", codes = list(create_component_settings("code",
      system = "RxNorm", code = "731370", display = "BNT162b2 vaccine"
    )))
  )))
))
str(eligible, max.level = 4)
```

## Guard: the state that loops on a condition

**Function:** `create_state_settings("Guard", allow = <Logic>, transition = <Transition>)`

A `Guard` is the one state type built specifically to hold a Logic tree (`allow`, required). It
has exactly one behavior: **re-test `allow` every simulated timestep, and don't advance past this
state until it passes.** That's a materially different thing from a `conditional_transition`,
which tests its condition once, the instant the state is reached, and immediately picks a branch
either way.

```{r guard-vs-conditional}
# Guard: blocks here, checking every timestep, until the person turns 18.
age_guard <- create_state_settings("Guard",
  allow = adult,
  transition = create_transition_settings("direct", to = "Enrollment")
)

# conditional_transition: evaluated once, right now -- routes immediately, doesn't wait.
age_branch <- create_state_settings("Simple",
  transition = create_transition_settings("conditional", options = list(
    list(condition = adult, transition = "Enrollment"),
    list(transition = "Ineligible") # condition omitted = the else branch
  ))
)
```

Use `Guard` when "not yet eligible" should mean *wait* (a person who's 16 now will enroll two
years from now, once `Age >= 18` starts passing). Use a conditional transition when "not eligible"
should mean *route elsewhere now* and never revisit the check.

## Worked example: a compound eligibility Guard

A trial population is often gated on a single `Age` check, but a stricter version would also
exclude specific active conditions with a `Not` + `ActiveCondition` combinator. Here's that
stricter gate actually built, combining everything above into one `Guard`:

```{r worked-example}
pregnancy_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "77386006", display = "Pregnancy (finding)"
)
vaccine_allergy_code <- create_component_settings("code",
  system = "RxNorm", code = "731370", display = "BNT162b2 vaccine"
)

trial_eligibility <- create_logic_settings("And", conditions = list(
  create_logic_settings("Age", operator = ">=", quantity = 16, unit = "years"),
  create_logic_settings("Not", condition = create_logic_settings("ActiveCondition",
    codes = list(pregnancy_code)
  )),
  create_logic_settings("Not", condition = create_logic_settings("ActiveAllergy",
    codes = list(vaccine_allergy_code)
  ))
))

states <- list(
  Initial = create_state_settings("Initial",
    transition = create_transition_settings("direct", to = "Eligibility Gate")
  ),
  "Eligibility Gate" = create_state_settings("Guard",
    allow = trial_eligibility,
    transition = create_transition_settings("direct", to = "Terminal")
  ),
  Terminal = create_state_settings("Terminal")
)

module_json <- build_module(name = "Logic and Guards Vignette Example", states = states)
cat(module_json)
```

The Guard's `allow` is exactly the R expression tree above, serialized: an `And` of an `Age`
check and two `Not`-wrapped `ActiveCondition`/`ActiveAllergy` checks -- and `build_module()`
validated the whole graph (one `Initial`, a `Terminal`, every transition resolves) the same way
it would for any other module.

## Where to go from here

This vignette covers `And`/`Or`/`Not`/`AtLeast`/`AtMost` and `Guard` in depth, plus three leaves
(`Age`, `Symptom`, `ActiveCondition`, `ActiveAllergy`) along the way. `vignette("logic-reference")`
is the complete reference for the other leaf predicates -- `Gender`, `Date`,
`SocioeconomicStatus`, `Race`, `Attribute`, `VitalSign`, `PriorState`, `ActiveMedication`,
`ActiveCarePlan`, `True`, and `False`.
