---
title: "build_disease_module() vs. build_module(): Recipe vs. Hand-Composed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{build_disease_module() vs. build_module(): Recipe vs. Hand-Composed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")`. Every other vignette so far has used
`build_module()` directly -- the primitive, generic entry point. This package also ships one
convenience function built entirely on top of it: `build_disease_module()`, a recipe for the
single most common module shape (onset -> diagnosis encounter -> optional resolution/death ->
`Terminal`). This vignette builds the *same* module both ways and proves they produce identical
output, so you can see exactly what the recipe buys you -- and exactly where its edges are.

```{r setup}
library(syntheaModuleR)
```

## The recipe

`example_generate.R` builds a minimal seasonal-rhinitis module with `build_disease_module()`
in one call:

```{r recipe}
condition_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "444316003", display = "Seasonal allergic rhinitis (disorder)")

recipe_module <- build_disease_module(
  name = "Recipe vs. Compose Example",
  condition_code = condition_code,
  diagnosis = "wellness",
  onset_delay = list(low = 1, high = 4, unit = "weeks"),
  resolves_after = list(low = 2, high = 6, unit = "weeks"),
  remarks = c("Built via build_disease_module().")
)

cat(jsonlite::toJSON(recipe_module, auto_unbox = TRUE, pretty = TRUE, null = "null"))
```

Note what you *didn't* have to do: no `create_state_settings()`, no `create_transition_settings()`,
no manual wiring of five state names to each other. `build_disease_module()` returns an R list
(`as_json = FALSE` always, unlike `build_module()`'s default), already validated -- hand it to
`write_module_json()` directly, or wrap it in `jsonlite::toJSON()` yourself as above just to look
at it.

## What it built, read from the inside

`build_disease_module()`'s own source (`module.R`) is not a separate code path -- it's
`create_state_settings()` and `create_transition_settings()` calls, same as any module you'd write
by hand, just decided programmatically from which optional arguments you passed. With
`onset_delay` and `resolves_after` given, `age_gate` and `death` omitted, and
`diagnosis = "wellness"`, it wires up exactly this chain:

```
Initial -> Onset Delay -> Condition Onset -> Diagnosis Encounter -> End Encounter
        -> Condition Resolves Delay -> Condition Resolves -> Terminal
```

- **`Onset Delay`** (`Delay`, `range` from `onset_delay`) -- present only because `onset_delay`
  was given; omit it and `Condition Onset` would follow `Initial` directly (or `Age Gate`, if
  `age_gate` were given instead -- that chain-building logic is what decides `Initial`'s target).
- **`Condition Onset`** (`ConditionOnset`, `codes = list(condition_code)`,
  `target_encounter = "Diagnosis Encounter"`).
- **`Diagnosis Encounter`** (`Encounter`, `wellness = TRUE` because `diagnosis = "wellness"` --
  it would be `encounter_class = <encounter_class>` for `diagnosis = "standalone"`).
- **`End Encounter`** (`EncounterEnd`) -- routes to `Condition Resolves Delay` because
  `resolves_after` was given (it would route straight to `Terminal` otherwise, or split
  probabilistically to a `Death` state via a `distributed_transition` if `death` were given
  instead).
- **`Condition Resolves Delay`** / **`Condition Resolves`** (`Delay` + `ConditionEnd`) -- present
  only because `resolves_after` was given; omitting it leaves the condition chronic.

## The same shape, hand-composed

Building that exact graph with `build_module()` + `create_state_settings()` means writing out
every state and transition the recipe generated for you:

```{r hand-composed}
composed_states <- list(
  "Onset Delay" = create_state_settings("Delay",
    range = create_component_settings("range", low = 1, high = 4, unit = "weeks"),
    transition = create_transition_settings("direct", to = "Condition Onset")),

  "Condition Onset" = create_state_settings("ConditionOnset",
    codes = list(condition_code),
    target_encounter = "Diagnosis Encounter",
    transition = create_transition_settings("direct", to = "Diagnosis Encounter")),

  "Diagnosis Encounter" = create_state_settings("Encounter",
    wellness = TRUE,
    transition = create_transition_settings("direct", to = "End Encounter")),

  "Condition Resolves Delay" = create_state_settings("Delay",
    range = create_component_settings("range", low = 2, high = 6, unit = "weeks"),
    transition = create_transition_settings("direct", to = "Condition Resolves")),

  "Condition Resolves" = create_state_settings("ConditionEnd",
    condition_onset = "Condition Onset",
    transition = create_transition_settings("direct", to = "Terminal")),

  "End Encounter" = create_state_settings("EncounterEnd",
    transition = create_transition_settings("direct", to = "Condition Resolves Delay")),

  Initial = create_state_settings("Initial",
    transition = create_transition_settings("direct", to = "Onset Delay")),

  Terminal = create_state_settings("Terminal")
)

composed_module <- build_module(
  name = "Recipe vs. Compose Example",
  states = composed_states,
  remarks = c("Built via build_disease_module()."),
  as_json = FALSE
)
```

The insertion order above -- `Onset Delay`, `Condition Onset`, `Diagnosis Encounter`,
`Condition Resolves Delay`, `Condition Resolves`, `End Encounter`, `Initial`, `Terminal` -- looks
like an odd way to write a `states` list by hand (`Initial` sixth-to-last, not first). It's written
in exactly the order `build_disease_module()`'s own source adds these states, for the identical
check below. In your own hand-composed modules, order the list however reads clearest to you --
Synthea resolves every state reference by *name*, never by position in the list, so state order
has no effect on how a module behaves. It only affects two things, both cosmetic: the key order in
pretty-printed JSON, and whether `identical()` on the R list says `TRUE`.

## Proof: identical output

Same `name`, same `remarks`, same `condition_code` object, same `gmf_version` default, same state
insertion order -- so the two module lists should be exactly, structurally identical, not just
"equivalent enough":

```{r identical-check}
identical(recipe_module, composed_module)
```

That's not a coincidence -- it's the point. `build_disease_module()` has no privileged access to
anything; it produces exactly what you'd write by hand for this one shape, in this one order, and
nothing more.

## Where the recipe's edges are

`build_disease_module()`'s parameter list -- `condition_code`, `diagnosis`, `encounter_class`,
`onset_delay`, `age_gate`, `resolves_after`, `death` -- covers one condition, one encounter, and
one linear path through it. The moment a module needs something that parameter list has no
argument for, the recipe can't express it at all -- not awkwardly, not with a workaround, just not
at all. Concretely, out of scope for `build_disease_module()`:

- More than one `ConditionOnset` in the same module, or any branching between them
- Any state type it doesn't call internally -- `Guard`, `MedicationOrder`, `Symptom`,
  `CarePlanStart`, `Vaccine`, `Procedure`, ... (see `vignette("layers")`'s full list of 31)
- `Logic` combinators for eligibility (`vignette("logic-and-guards")`) -- the recipe's only
  built-in gate is a single flat `Age` check via `age_gate`
- `distributed_transition`/`conditional_transition` branching beyond the one death/no-death split
  it already knows how to build
- `SetAttribute`, `Counter`, or anything involving person attributes at all
  (`vignette("set-attribute-precedence")`)

Extending the hand-composed version past what the recipe could ever express is just adding one
more state -- e.g. prescribing an antihistamine during the encounter, something
`build_disease_module()` has no parameter for whatsoever:

```{r extended}
extended_states <- composed_states
extended_states[["Diagnosis Encounter"]] <- create_state_settings("Encounter",
  wellness = TRUE,
  transition = create_transition_settings("direct", to = "Prescribe Antihistamine"))
extended_states[["Prescribe Antihistamine"]] <- create_state_settings("MedicationOrder",
  codes = list(create_component_settings("code",
    system = "RxNorm", code = "997219", display = "Cetirizine hydrochloride 10 MG Oral Tablet")),
  reason = "Condition Onset",
  transition = create_transition_settings("direct", to = "End Encounter"))

extended_module <- build_module(name = "Extended Example", states = extended_states)
```

That call succeeds because `build_module()` never cared how its `states` list was assembled --
by a recipe, by hand, or (as here) by taking a recipe's output and grafting one more state onto
it.

## When to reach for which

Start with `build_disease_module()` when what you're building genuinely is "one condition, one
encounter, optionally gated by age, optionally resolves or kills" -- it's less to type and
produces the identical thing you'd have written anyway. Drop to `build_module()` the moment the
shape stops fitting: multiple conditions, non-linear branching, any state type outside
`ConditionOnset`/`Encounter`/`EncounterEnd`/`ConditionEnd`/`Delay`/`Guard`/`Death`, or anything
involving `Logic` beyond a single `Age` gate. `example_compose.R` and
`example_trial_emulation.R` are both entirely in `build_module()` territory -- neither
fits the recipe's shape at all.
