---
title: "SetAttribute: Value Source Precedence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SetAttribute: Value Source Precedence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette assumes you've read `vignette("layers")` and, for the `distribution` examples
below, `vignette("distribution")`. It covers one state type in isolation because it's the one
place in this package where a real constraint from the Java engine is **not** enforced at build
time -- worth understanding on its own, not just as a footnote.

```{r setup}
library(syntheaModuleR)
```

## Seven ways to give a value, one attribute set

**Function:** `create_state_settings("SetAttribute", attribute = ..., ..., transition = ...)`

`SetAttribute` writes a value onto a named person attribute. `attribute` and `transition` are the
only required fields; everything else is a *value source*, and there are seven of them:
`value`, `value_code`, `value_attribute`, `range`, `expression`, `series_data`, `distribution`.
(`period` is an eighth optional field, paired with `series_data` -- see below.)

Every other multi-source field in this package that has an "only one of these makes sense"
relationship is enforced with a `one_of` constraint -- `Delay`'s `range`/`exact`/`distribution`,
`VitalSign`'s four value sources, `Procedure`'s `duration`/`distribution` (see
`vignette("distribution")`). `SetAttribute` is the one exception: this package's schema for it has
**no `one_of` group at all**. You can pass all seven at once and `create_state_settings()` will
build it without complaint.

That's not an oversight -- it mirrors the Java engine itself, which doesn't reject multiple value
sources either. Instead it resolves them by **precedence**, checking each in a fixed order and
using the first one present:

```
expression  >  range  >  series_data  >  distribution  >  value_code  >  value_attribute  >  value
```

If you pass only one source, precedence is invisible -- it just works. It only matters once more
than one is present, which this package will silently let you do.

## Each source on its own

```{r value}
create_state_settings("SetAttribute",
  attribute = "risk_score", value = 0,
  transition = create_transition_settings("direct", to = "Next")
)
```

`value_code` assigns a `Code` component rather than a scalar -- useful for an attribute that
represents a coded concept, like a blood type:

```{r value_code}
create_state_settings("SetAttribute",
  attribute = "blood_type",
  value_code = create_component_settings("code",
    system = "SNOMED-CT", code = "112144000", display = "Blood group A Rh(D) positive"
  ),
  transition = create_transition_settings("direct", to = "Next")
)
```

`value_attribute` copies another attribute's current value:

```{r value_attribute}
create_state_settings("SetAttribute",
  attribute = "risk_score_snapshot", value_attribute = "risk_score",
  transition = create_transition_settings("direct", to = "Next")
)
```

`range` samples uniformly between `low` and `high` -- the same `range` component from
`vignette("layers")`:

```{r range}
create_state_settings("SetAttribute",
  attribute = "risk_score",
  range = create_component_settings("range", low = 0, high = 100),
  transition = create_transition_settings("direct", to = "Next")
)
```

`distribution` samples from any of the five `Distribution` kinds covered in
`vignette("distribution")`:

```{r distribution}
create_state_settings("SetAttribute",
  attribute = "risk_score",
  distribution = create_component_settings("distribution",
    kind = "GAUSSIAN",
    mean = 50, standardDeviation = 15, min = 0, max = 100
  ),
  transition = create_transition_settings("direct", to = "Next")
)
```

`expression` evaluates a small formula string against the person's current attributes --
Synthea's own expression mini-language (see the GMF wiki), not something this package parses or
validates; it's passed through as a raw string:

```{r expression}
create_state_settings("SetAttribute",
  attribute = "risk_score", expression = "#{risk_score} + 5",
  transition = create_transition_settings("direct", to = "Next")
)
```

`series_data` (paired with the optional `period`) drives a time-varying attribute through a
sequence of values rather than a single sample. Like `expression`, this package doesn't have a
dedicated component for its shape -- there's no `array` rule applied and no schema check beyond
"it's present" -- so it's passed through as a raw R list, the same way `AllergyOnset`'s
`reactions` or `ImagingStudy`'s `series` items are (see `vignette("layers")`'s "out of scope"
note). Consult the GMF wiki for its exact expected shape before relying on it.

## The gotcha, live

Because there's no `one_of` here, this builds without error -- and without a warning that one of
the two fields you gave is now dead:

```{r two-sources-no-error}
conflicting <- create_state_settings("SetAttribute",
  attribute = "risk_score",
  range = create_component_settings("range", low = 0, high = 100),
  distribution = create_component_settings("distribution",
    kind = "GAUSSIAN",
    mean = 50, standardDeviation = 15
  ),
  transition = create_transition_settings("direct", to = "Next")
)
names(conflicting)
```

Per the precedence order above, `range` outranks `distribution` -- so at runtime, in the real
Synthea engine, the `distribution` field in this state is inert. Nothing in this package's
`validate_module()` will catch that either, since graph-level validation checks structure (states
exist, transitions resolve), not which combinations of optional fields are meaningful together.
**The discipline of passing exactly one value source per `SetAttribute` call is on you, not the
toolkit.**

## Worked example: baseline, derive, copy

```{r worked-example}
states <- list(
  Initial = create_state_settings("Initial",
    transition = create_transition_settings("direct", to = "Set Baseline Risk")
  ),

  # 1. range: a random baseline value, sampled once.
  "Set Baseline Risk" = create_state_settings("SetAttribute",
    attribute = "risk_score",
    range = create_component_settings("range", low = 0, high = 50),
    transition = create_transition_settings("direct", to = "Apply Risk Modifier")
  ),

  # 2. expression: derive a new value from the attribute set in step 1.
  "Apply Risk Modifier" = create_state_settings("SetAttribute",
    attribute = "risk_score", expression = "#{risk_score} + 10",
    transition = create_transition_settings("direct", to = "Snapshot Risk")
  ),

  # 3. value_attribute: copy the derived value somewhere else, verbatim.
  "Snapshot Risk" = create_state_settings("SetAttribute",
    attribute = "risk_score_at_enrollment", value_attribute = "risk_score",
    transition = create_transition_settings("direct", to = "Terminal")
  ),
  Terminal = create_state_settings("Terminal")
)

module_json <- build_module(name = "SetAttribute Precedence Vignette Example", states = states)
cat(module_json)
```

Each of these three `SetAttribute` states uses exactly one value source, so precedence never
actually has to break a tie here -- which is the discipline the previous section argued for:
one source per call, chosen deliberately, rather than relying on the precedence order to pick a
winner for you.

## Where to go from here

A side-by-side of `build_disease_module()`'s onset -> encounter -> resolve/death recipe against
hand-composing the same shape with `build_module()` directly is still open, and would be a natural
next vignette -- it would also show `SetAttribute` used for real inside a recipe function rather
than in isolation.
