---
title: "State Reference: All 31 State Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State Reference: All 31 State Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`vignette("layers")` introduces `create_state_settings()` with two state types
(`ConditionOnset`, `Guard`); a handful of others get their own worked examples in
`vignette("logic-and-guards")`, `vignette("distribution")`, and
`vignette("set-attribute-precedence")`. This vignette is the complete reference for all 31 --
most of which don't otherwise appear in any vignette at all.

```{r setup}
library(syntheaModuleR)

# Local convenience for this vignette only -- not a package function.
direct <- function(to) create_transition_settings("direct", to = to)
```

## Quick reference

`transition` column: **req** = required, **forbid** = forbidden (`Terminal` only), **opt** =
optional (`Observation`, only when nested inside `MultiObservation`/`DiagnosticReport`).

| `type`            | Required                          | Array fields         | `transition` |
|--------------------|-------------------------------------|------------------------|---------------|
| `Initial`          |                                      |                        | req |
| `Terminal`         |                                      |                        | forbid |
| `Simple`           |                                      |                        | req |
| `CallSubmodule`    | `submodule`                          |                        | req |
| `Physiology`       | `model`, `solver`, `step_size`, `sim_duration`, `alt_direct_transition` | `inputs`, `outputs` | req |
| `Guard`            | `allow`                              |                        | req |
| `Delay`            | (one of `range`/`exact`/`distribution`) |                    | req |
| `SetAttribute`     | `attribute`                          |                        | req |
| `Counter`          | `attribute`, `action`                |                        | req |
| `Encounter`        |                                      | `codes`                | req |
| `EncounterEnd`     |                                      |                        | req |
| `ConditionOnset`   | `codes`                              | `codes`                | req |
| `ConditionEnd`     |                                      | `codes`                | req |
| `AllergyOnset`     | `codes`                              | `codes`, `reactions`   | req |
| `AllergyEnd`       |                                      | `codes`                | req |
| `MedicationOrder`  | `codes`                              | `codes`                | req |
| `MedicationEnd`    |                                      | `codes`                | req |
| `CarePlanStart`    | `codes`                              | `codes`, `activities`, `goals` | req |
| `CarePlanEnd`      |                                      | `codes`                | req |
| `Procedure`        | `codes`                              | `codes`                | req |
| `VitalSign`        | `vital_sign` (+ one of `exact`/`range`/`expression`/`distribution`) |  | req |
| `Observation`      | `codes`                              | `codes`                | opt |
| `MultiObservation` | `codes`, `observations`              | `codes`, `observations`| req |
| `DiagnosticReport` | `codes`, `observations`              | `codes`, `observations`| req |
| `ImagingStudy`     | `procedure_code`, `series`           | `series`                | req |
| `Symptom`          | `symptom`                             |                        | req |
| `Device`           | `code` (singular)                    |                        | req |
| `DeviceEnd`        |                                      | `codes` (plural)        | req |
| `SupplyList`       | `supplies`                            | `supplies`              | req |
| `Death`            |                                      | `codes`                | req |
| `Vaccine`          | `series` (dose number), `codes`      | `codes`                | req |

A few reusable codes for the examples below:

```{r shared-codes}
strep_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "43878008", display = "Streptococcal sore throat (disorder)"
)
penicillin_code <- create_component_settings("code",
  system = "RxNorm", code = "7980", display = "Penicillin"
)
acetaminophen_code <- create_component_settings("code",
  system = "RxNorm", code = "313782", display = "Acetaminophen 325 MG Oral Tablet"
)
diabetes_careplan_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "698360004", display = "Diabetes self management plan"
)
```

## Control flow

`Initial`, `Terminal`, and `Guard` are recapped from `vignette("layers")`; `Delay` from
`vignette("distribution")`. `Simple` is a plain pass-through node -- no fields, just a transition
-- used throughout `example_trial_emulation.R` as a routing point between branches.
`CallSubmodule` hands control to another module file by path; this package doesn't check that
the referenced submodule actually exists in the target Synthea checkout.

```{r control-flow}
create_state_settings("Initial", transition = direct("Next"))
create_state_settings("Terminal")
create_state_settings("Simple", transition = direct("Next"))
create_state_settings("CallSubmodule",
  submodule = "medications/otc_antihistamine",
  transition = direct("Next")
)
create_state_settings("Guard", allow = create_logic_settings("True"), transition = direct("Next"))
create_state_settings("Delay",
  exact = create_component_settings("exact", quantity = 7, unit = "days"),
  transition = direct("Next")
)
```

`Physiology` is gated behind the `physiology.state.enabled` config flag (off by default), which
is why `alt_direct_transition` is required -- it's the fallback path used whenever that flag is
off. `model` and `solver` are free strings this package doesn't restrict:

```{r physiology}
create_state_settings("Physiology",
  model = "BloodPressureModel.xml", solver = "rk4", step_size = 0.1, sim_duration = 1,
  alt_direct_transition = "Next",
  transition = direct("Next")
)
```

## Attributes

`SetAttribute` gets its own vignette (`vignette("set-attribute-precedence")`) -- recapped here
for completeness. `Counter` increments or decrements a numeric attribute by `amount` (default 1):

```{r attributes}
create_state_settings("SetAttribute",
  attribute = "risk_score", value = 0,
  transition = direct("Next")
)
create_state_settings("Counter",
  attribute = "dose_count", action = "increment",
  transition = direct("Next")
)
```

## Encounters

```{r encounters}
create_state_settings("Encounter", wellness = TRUE, transition = direct("Next"))
create_state_settings("EncounterEnd",
  discharge_disposition = create_component_settings("code",
    system = "SNOMED-CT", code = "306689006", display = "Discharge to home"
  ),
  transition = direct("Next")
)
```

## Conditions

`ConditionOnset` is recapped from `vignette("layers")`:

```{r conditions}
create_state_settings("ConditionOnset",
  codes = list(strep_code), target_encounter = "Diagnosis Encounter",
  transition = direct("Diagnosis Encounter")
)
create_state_settings("ConditionEnd",
  condition_onset = "Condition Onset",
  transition = direct("Next")
)
```

## Allergies

`allergy_type`/`category` are free strings (e.g. `"allergy"`/`"intolerance"`,
`"food"`/`"medication"`/`"environment"`), not restricted by this package:

```{r allergies}
create_state_settings("AllergyOnset",
  codes = list(penicillin_code),
  allergy_type = "allergy", category = "medication",
  transition = direct("Next")
)
create_state_settings("AllergyEnd",
  allergy_onset = "Allergy Onset",
  transition = direct("Next")
)
```

## Medications

`MedicationOrder`'s `reason` resolves against a person attribute, a prior state's onset name, or
a `ConditionOnset` state -- see `vignette("layers")`'s note on this:

```{r medications}
create_state_settings("MedicationOrder",
  codes = list(acetaminophen_code),
  reason = "Condition Onset", transition = direct("Next")
)
create_state_settings("MedicationEnd",
  medication_order = "Prescribe Medication",
  transition = direct("Next")
)
```

## Care plans

`activities` and `goals` are both lists of `code` components, same as `codes`:

```{r careplans}
create_state_settings("CarePlanStart",
  codes = list(diabetes_careplan_code),
  activities = list(create_component_settings("code",
    system = "SNOMED-CT", code = "160670007", display = "Diabetic diet"
  )),
  goals = list(create_component_settings("code",
    system = "SNOMED-CT", code = "160688004", display = "Weight reduction"
  )),
  transition = direct("Next")
)
create_state_settings("CarePlanEnd", careplan = "Start Care Plan", transition = direct("Next"))
```

## Procedures, symptoms, vitals, death, vaccination

`vignette("distribution")` covers `VitalSign` in depth; recapped here. `Symptom` tracks severity
1-100 -- the highest value across all causes drives urgency logic elsewhere in Synthea:

```{r clinical-events}
create_state_settings("Procedure",
  codes = list(strep_code),
  duration = create_component_settings("exact", quantity = 30, unit = "minutes"),
  transition = direct("Next")
)
create_state_settings("Symptom",
  symptom = "Chest Pain",
  range = create_component_settings("range", low = 10, high = 40),
  transition = direct("Next")
)
create_state_settings("VitalSign",
  vital_sign = "BMI", unit = "kg/m2",
  exact = create_component_settings("exact", quantity = 24), transition = direct("Next")
)
```

Omitting both `range` and `exact` on `Death` means immediate death; giving one delays it:

```{r death}
create_state_settings("Death", condition_onset = "Condition Onset", transition = direct("Next"))
```

`Vaccine`'s `series` is the dose number (an integer) -- not to be confused with `ImagingStudy`'s
`series` list below, same field name on two unrelated state types:

```{r vaccine}
flu_vaccine_code <- create_component_settings("code",
  system = "CVX", code = "141", display = "Influenza, seasonal, injectable"
)
create_state_settings("Vaccine",
  series = 1, codes = list(flu_vaccine_code),
  transition = direct("Next")
)
```

## Observations and imaging

A standalone `Observation` (used as its own state, so `transition` applies normally):

```{r observation}
temperature_code <- create_component_settings("code",
  system = "LOINC", code = "8310-5", display = "Body temperature"
)
create_state_settings("Observation",
  codes = list(temperature_code),
  exact = create_component_settings("exact", quantity = 98.6, unit = "degF"), unit = "degF",
  category = "vital-signs", transition = direct("Next")
)
```

`MultiObservation` and `DiagnosticReport` both take a list of nested `Observation`-shaped
settings via `observations` -- built the same way, just with `transition = NULL` (the one case
where `Observation`'s transition is optional rather than required, per `vignette("layers")`):

```{r multiobservation}
systolic_code <- create_component_settings("code",
  system = "LOINC", code = "8480-6", display = "Systolic blood pressure"
)
diastolic_code <- create_component_settings("code",
  system = "LOINC", code = "8462-4", display = "Diastolic blood pressure"
)

bp_panel_code <- create_component_settings("code",
  system = "LOINC", code = "35094-2", display = "Blood pressure panel"
)

create_state_settings("MultiObservation",
  codes = list(bp_panel_code),
  observations = list(
    create_state_settings("Observation",
      codes = list(systolic_code),
      exact = create_component_settings("exact", quantity = 120), unit = "mmHg"
    ),
    create_state_settings("Observation",
      codes = list(diastolic_code),
      exact = create_component_settings("exact", quantity = 80), unit = "mmHg"
    )
  ),
  transition = direct("Next")
)
```

`DiagnosticReport` is the same shape as `MultiObservation`, just without the optional `category`
field:

```{r diagnosticreport}
create_state_settings("DiagnosticReport",
  codes = list(bp_panel_code),
  observations = list(
    create_state_settings("Observation",
      codes = list(systolic_code),
      exact = create_component_settings("exact", quantity = 120), unit = "mmHg"
    )
  ),
  transition = direct("Next")
)
```

`ImagingStudy`'s `series` items follow `HealthRecord.ImagingStudy.Series`'s shape
(`modality`, `body_site`, `instances`) -- this package passes them through as raw lists, not
independently validated:

```{r imagingstudy}
xray_procedure_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "399208008", display = "Plain chest X-ray"
)

create_state_settings("ImagingStudy",
  procedure_code = xray_procedure_code,
  series = list(list(
    modality = create_component_settings("code",
      system = "DICOM-DCM", code = "DX", display = "Digital Radiography"
    ),
    body_site = create_component_settings("code",
      system = "SNOMED-CT", code = "51185008", display = "Thoracic structure"
    ),
    instances = list(list(
      title = "Chest X-ray, PA view",
      sop_class = create_component_settings("code",
        system = "DICOM-SOP", code = "1.2.840.10008.5.1.4.1.1.1.1",
        display = "Digital X-Ray Image Storage"
      )
    ))
  )),
  transition = direct("Next")
)
```

## Devices and supplies

`Device.code` is **singular** (one `Code`) -- `DeviceEnd.codes` is **plural** (a list), the
asymmetry `vignette("layers")` calls out explicitly:

```{r devices}
pacemaker_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "14106009", display = "Cardiac pacemaker"
)

create_state_settings("Device",
  code = pacemaker_code,
  manufacturer = "Acme Medical", model = "X100", transition = direct("Next")
)
create_state_settings("DeviceEnd", device = "Implant Device", transition = direct("Next"))
```

Each `supplies` item is a raw `list(code = <a code component>, quantity = <n>)`, not
independently validated:

```{r supplylist}
gauze_code <- create_component_settings("code",
  system = "SNOMED-CT", code = "705076002", display = "Sterile gauze dressing"
)

create_state_settings("SupplyList",
  supplies = list(list(code = gauze_code, quantity = 10)),
  transition = direct("Next")
)
```

## Where to go from here

Most of the 31 examples above are standalone -- built and printed, not assembled into a full
module, since wiring 31 states into one coherent graph would obscure rather than clarify each
one's shape. `vignette("layers")` and `example_compose.R`/`example_trial_emulation.R`
show several of these state types wired together into real, validated modules.
