---
title: "Transition Reference: All Six Kinds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transition Reference: All Six Kinds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`vignette("layers")` introduces `create_transition_settings()` and shows three of its six kinds
(`direct`, `conditional`, `distributed`). This vignette is the complete reference, including the
three (`complex`, `lookup_table`, `type_of_care`) that don't otherwise appear in any vignette.

```{r setup}
library(syntheaModuleR)
```

## Quick reference

| `kind`         | Required                                      | Notes |
|----------------|------------------------------------------------|-------|
| `direct`       | `to`                                            | always goes to the same state |
| `distributed`  | `options` (`transition`, `distribution`)        | picks one option at random |
| `conditional`  | `options` (`condition` [optional on the last], `transition`) | first match wins; last option with no `condition` is the else branch |
| `complex`      | `options` (`condition` optional, plus exactly one of `transition`/`distributions`) | combines conditional and distributed routing |
| `lookup_table` | `lookup_table_name`, `options` (`transition`, `default_probability`) | reads probabilities from an external CSV this package doesn't generate |
| `type_of_care` | `ambulatory`, `telemedicine`, `emergency`       | routes by insurance/telemedicine-adoption date, config-driven |

## `direct`, `conditional`, `distributed` (recap)

```{r recap}
create_transition_settings("direct", to = "Diagnosis Encounter")

create_transition_settings("conditional", options = list(
  list(
    condition = create_logic_settings("Age", operator = ">=", quantity = 18, unit = "years"),
    transition = "Adult Dosing"
  ),
  list(transition = "Pediatric Dosing")
))

create_transition_settings("distributed", options = list(
  list(transition = "Resolves", distribution = 0.9),
  list(transition = "Death", distribution = 0.1)
))
```

## `complex`

Each option is gated by an optional `condition` and is itself either a direct `transition` or a
nested `distributions` list -- so a single transition can mix "route by condition" and "route by
probability" in one place. Here, female patients always go to one pathway; everyone else splits
50/50 between two others:

```{r complex}
create_transition_settings("complex", options = list(
  list(
    condition = create_logic_settings("Gender", gender = "F"),
    transition = "Female Pathway"
  ),
  list(distributions = list(
    list(transition = "Pathway A", distribution = 0.5),
    list(transition = "Pathway B", distribution = 0.5)
  ))
))
```

Each option needs *exactly one* of `transition` or `distributions` -- giving both, or neither, is
rejected the same way every other `one_of` constraint in this package is:

```{r complex-error, error = TRUE}
create_transition_settings("complex", options = list(
  list(transition = "A", distributions = list(list(transition = "B", distribution = 1)))
))
```

## `lookup_table`

Reads transition probabilities from a CSV keyed by person attributes (plus special `age`/`time`
range columns), under Synthea's `generate.lookup_tables` config directory. This package only
wires the reference -- the CSV itself has to already exist in the target Synthea checkout:

```{r lookup_table}
create_transition_settings("lookup_table",
  lookup_table_name = "risk_by_demographics.csv",
  options = list(
    list(transition = "High Risk Path", default_probability = 0.3),
    list(transition = "Low Risk Path", default_probability = 0.7)
  )
)
```

## `type_of_care`

Routes based on insurance and the simulated date relative to a configured telemedicine-adoption
point (modeling the COVID-era telehealth shift). Entirely config-driven -- this package just
wires the three state-name targets:

```{r type_of_care}
create_transition_settings("type_of_care",
  ambulatory = "Ambulatory Visit", telemedicine = "Telemedicine Visit", emergency = "ER Visit"
)
```
